<!--
  SHARED BODY for the Exam 1 review problems.

  This file is NOT a page. It is included as a child document by both
  exam1_review_problems.Rmd and exam1_review_problems_solns.Rmd, each of which
  sets the SOLUTIONS flag before including it. Writing the questions once means
  an answer can never drift away from the question it belongs to -- the two
  pages were previously ~300 lines of hand-synced near-duplicate.

  The leading underscore keeps docs/build_site.R from rendering it as a page of
  its own (see the `^[^_]` pattern in its subdirectory loop).

  To add a problem: write the question once, and wrap every answer in ans() for
  short fill-ins, or an `eval = SOLUTIONS` chunk for worked steps. Compute the
  answer from the same vector the question prints, and stopifnot() anything you
  also state in prose.
-->

```{r review-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggthemes)
library(ggpubr)

stopifnot(exists("SOLUTIONS"), is.logical(SOLUTIONS), length(SOLUTIONS) == 1L)

## Short inline answer. Renders a blank in the problems copy and red text in the
## solutions copy -- the same red the Spring 2024 solutions used.
##
## The red comes from a CLASS, not an inline style, so that one @media print rule
## below can keep it red in the PDF. Chrome's default print-color-adjust is
## "economy", which recolours body text to black to save ink -- inline styles do
## not survive it either, so the class is what makes the fix possible.
ans <- function(x, blank = "$\\_\\_\\_\\_\\_\\_\\_\\_$") {
  if (SOLUTIONS) sprintf('<span class="soln">%s</span>', x) else blank
}

## A whole answer paragraph, for parts that want a sentence rather than a value.
## Emits nothing at all in the problems copy.
why <- function(x) {
  if (SOLUTIONS) cat(sprintf('\n\n<span class="soln">%s</span>\n\n', x))
}

## Vertical space for students to work in, on the problems copy only.
workspace <- function(lines = 4) {
  if (!SOLUTIONS) cat(strrep("<br>\n", lines))
}

## Standard table look, matching the rest of the site.
show_tbl <- function(d, ...) {
  kable(d, format = "html", row.names = FALSE, escape = FALSE, ...) %>%
    kable_classic(full_width = FALSE, html_font = "Cambria") %>%
    kable_styling(bootstrap_options = "striped")
}

## A cell that is blank for students and filled in red for the key. Vectorised so
## a whole column can be built at once.
cellv <- function(x) {
  if (SOLUTIONS) sprintf('<span class="soln">%s</span>', x) else ""
}

## A blank cell that reserves room to WRITE IN, for the tables whose whole point
## is that the student works the formula out in the cell. An empty <td> collapses
## to a single line of height, which leaves nowhere to put a fraction; the div
## props the row open instead. Solutions copy is unaffected -- the answer fills
## the cell there.
cellbox <- function(x, height = "1.1in") {
  if (SOLUTIONS) sprintf('<span class="soln">%s</span>', x)
  else sprintf('<div style="min-height:%s"></div>', height)
}

## A column that is only PARTLY blank: the entries at `blanks` are left for the
## student (and shown in red in the key), every other entry is printed in both
## copies. Used where the point of the exercise is that the student recovers the
## missing entries from the ones that are there -- the frequencies sum to n, the
## relative frequencies sum to 1, and each CRF is the running total of the RFs.
cellsome <- function(x, blanks) {
  out <- as.character(x)
  out[blanks] <- cellv(out[blanks])
  out
}

## Print a set of observations in the notation used in the lecture notes:
##   X = {2, 5, 6, 9, 13}
## Character values are wrapped in \text{} so they typeset upright rather than as
## italic maths variables. Every problem that hands students raw data uses this,
## so the review set reads the same way the notes do.
## Long sets have to be broken by hand: display maths does not wrap, so a set of
## 20 categories on one line runs off the right edge of the page and is simply
## cut off in the PDF. Anything longer than per_line values is split across a
## `gathered` environment instead. Character values get fewer per line because
## the words are much wider than digits.
show_set <- function(v, name = "X",
                     per_line = if (is.character(v)) 6L else 10L) {
  vals <- if (is.character(v)) sprintf("\\text{%s}", v) else format(v, trim = TRUE)

  if (length(vals) <= per_line) {
    cat(sprintf("\n\n$$%s = \\{ %s \\}$$\n\n", name, paste(vals, collapse = ", \\ ")))
    return(invisible(NULL))
  }

  rows  <- split(vals, ceiling(seq_along(vals) / per_line))
  lines <- vapply(rows, paste, character(1), collapse = ", \\ ")
  ## Trailing comma on every line but the last, so the break reads as a
  ## continuation rather than the end of the set.
  lines[-length(lines)] <- paste0(lines[-length(lines)], ",")
  cat(sprintf("\n\n$$\\begin{gathered}\n%s = \\{ %s \\}\n\\end{gathered}$$\n\n",
              name, paste(lines, collapse = " \\\\\n")))
}

## The class's quartile rule: median of halves, dropping the median itself when n
## is odd. Used by every problem that needs a quartile, so no problem can quietly
## disagree with Q16 of the exam or with the Week 3 notes.
q_halves <- function(x) {
  x <- sort(x); n <- length(x); h <- n %/% 2
  c(Q1 = median(x[1:h]), Q2 = median(x), Q3 = median(x[(n - h + 1):n]))
}
```

```{css, echo=FALSE}
/* Answers in the solutions copy. */
.soln { color: #c00000; }

/* These pages are published as HTML for the website AND printed to PDF for
   Canvas (see docs/build_review.R). Everything below is for the PDF. */
@media print {
  /* Chrome's default print-color-adjust is "economy", which recolours body text
     to black to save ink -- so without this the red answers print black and the
     solutions copy becomes indistinguishable from the problems copy. */
  .soln {
    color: #c00000 !important;
    -webkit-print-color-adjust: exact;
    print-color-adjust: exact;
  }

  /* Keep a problem's heading with the problem, and never split a table or a
     figure across a page break. */
  h2 { break-after: avoid; page-break-after: avoid; }
  table, div.figure, p > img { break-inside: avoid; page-break-inside: avoid; }
}
```

```{r review-intro, results='asis'}
if (SOLUTIONS) {
  cat('Worked solutions. Answers are shown <span style="color: red;">in red</span>.\n\n')
}
```

These problems cover everything on Exam 1: Weeks 1 through 3, up to and including
cumulative distributions. They are **more computational than the exam** by design. The exam
asks you to interpret values, identify equations and read plots; these ask you to work the
arithmetic behind them.

Work them by hand. Every dataset is small enough to do on paper, and a calculator is fine.

---

## Problem 1 — The anatomy of data

A researcher recorded the following information for each of six trees in a city park.

```{r p01-data}
trees <- data.frame(
  `Tree ID`           = c("T-01", "T-02", "T-03", "T-04", "T-05", "T-06"),
  `Species`           = c("Maple", "Oak", "Maple", "Pine", "Oak", "Maple"),
  `Health rating`     = c("Good", "Fair", "Excellent", "Good", "Poor", "Good"),
  `Branches removed`  = c(3, 0, 5, 1, 8, 2),
  `Trunk width (cm)`  = sprintf("%.1f", c(31.4, 58.2, 22.7, 44.0, 61.9, 27.5)),
  check.names = FALSE, stringsAsFactors = FALSE
)
show_tbl(trees)
```

**(a)** How many observations are in this data table, and how many variables?

`r ans("6 observations and 5 variables")`

**(b)** What is the value of $n$ for this sample?

`r ans("$n = 6$")`

**(c)** Classify each of the four measured variables as nominal, ordinal, discrete or
continuous.

```{r p01-types}
show_tbl(data.frame(
  Variable = c("Species", "Health rating", "Branches removed", "Trunk width (cm)"),
  `Your answer` = cellv(c("nominal", "ordinal", "discrete", "continuous")),
  check.names = FALSE))
```

```{r p01-why, results='asis'}
why(paste("<b>Species</b> is a category with no natural ordering, so nominal.",
          "<b>Health rating</b> is a category <i>with</i> an ordering -- poor, fair, good,",
          "excellent -- so ordinal. <b>Branches removed</b> counts things, so it takes",
          "whole-number values only: discrete. <b>Trunk width</b> is a measurement on a",
          "scale, recorded to a tenth of a centimetre, so continuous.",
          "<br><br>Note that <i>Tree ID</i> is a label, not a measured variable -- it",
          "identifies the observation rather than describing it."))
```

**(d)** These six trees are a sample. Describe the population the researcher is presumably
interested in, and say whether $N$ is known.

```{r p01-pop, results='asis'}
why(paste("The population is all of the trees in the city park (or, more broadly, all",
          "trees the park service manages). $N$ is the number of trees in that",
          "population. Here it is not given -- and unlike most populations we will meet,",
          "this one is finite and could actually be counted."))
workspace(3)
```

---

## Problem 2 — Statistics, parameters, and inference

A city has 8,500 registered bicycles. A researcher takes a random sample of 200 of them and
finds that 46 have no working lights. She also measures the age of each sampled bicycle and
finds a mean age of 6.2 years.

```{r p02-calc}
N_bikes <- 8500; n_bikes <- 200; no_lights <- 46
phat <- no_lights / n_bikes
stopifnot(phat == 0.23)
```

**(a)** Give the values of $N$ and $n$.

`r ans("$N = 8{,}500$ and $n = 200$")`

**(b)** Show how to compute the sample proportion of bicycles with no working lights. Which symbol
denotes it?

`r ans("$\\hat{p} = \\frac{\\text{number with no working lights}}{n} = \\frac{46}{200} = 0.23$")`

```{r p02-b, results='asis'}
why(paste("The <b>sample</b> proportion is written $\\hat{p}$. The hat marks it as a",
          "quantity computed from a sample. The corresponding population proportion --",
          "the fraction of all 8,500 bicycles with no working lights -- is written $p$",
          "and is unknown."))
workspace(2)
```

**(c)** For each quantity, say whether it is a **statistic** or a **parameter**, and give
the symbol that denotes it.

```{r p02-symbols}
show_tbl(data.frame(
  Quantity = c("The mean age of the 200 sampled bicycles",
               "The mean age of all 8,500 registered bicycles",
               "The proportion of the 200 sampled with no lights",
               "The proportion of all 8,500 with no lights"),
  `Statistic or parameter` = cellv(c("statistic", "parameter", "statistic", "parameter")),
  Symbol = cellv(c("$\\bar{x}$", "$\\mu$", "$\\hat{p}$", "$p$")),
  check.names = FALSE))
```

**(d)** The researcher reports the mean age of the 200 bicycles she measured, then uses it
to estimate the mean age of all 8,500. Which of those two acts is descriptive statistics
and which is inferential? Explain what makes the difference.

```{r p02-d, results='asis'}
why(paste("Reporting the mean of the 200 measured bicycles is <b>descriptive</b>: it",
          "summarizes the data in hand and claims nothing beyond it. Using that mean to",
          "estimate the mean age of all 8,500 is <b>inferential</b>: it reaches past the",
          "sample to a population that was never fully observed.",
          "<br><br>The difference is not which data you have -- both acts use the same",
          "200 bicycles. It is <i>what you are claiming</i>: whether the conclusion stops",
          "at the sample or extends beyond it."))
workspace(4)
```

---

## Problem 3 — Reading the formulas

On the exam the formulas are provided **unlabeled** — the equations only, with no names.
This problem is practice at telling them apart.

**(a)** Each equation below computes one of the quantities in the list. Write the name of
the quantity next to its equation.

> sample mean &nbsp;&middot;&nbsp; population mean &nbsp;&middot;&nbsp; sample proportion
> &nbsp;&middot;&nbsp; population proportion &nbsp;&middot;&nbsp; sample variance
> &nbsp;&middot;&nbsp; sample standard deviation &nbsp;&middot;&nbsp; interquartile range

```{r p03-match}
show_tbl(data.frame(
  Equation = c("$\\dfrac{1}{n}\\sum_{i=1}^{n} x_i$",
               "$\\sqrt{\\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}}$",
               "$Q3 - Q1$",
               "$\\dfrac{1}{N}\\sum_{i=1}^{N} x_i$",
               "$\\dfrac{\\text{number of observations in the category}}{n}$",
               "$\\dfrac{\\text{number of observations in the category}}{N}$",
               "$\\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}$"),
  `What it computes` = cellv(c("sample mean", "sample standard deviation",
                               "interquartile range", "population mean",
                               "sample proportion", "population proportion",
                               "sample variance")),
  Symbol = cellv(c("$\\bar{x}$", "$s$", "$IQR$", "$\\mu$", "$\\hat{p}$", "$p$", "$s^2$")),
  check.names = FALSE))
```

```{r p03-a, results='asis'}
why(paste("Three things give each one away. The <b>divisor</b>: $n$ for a sample mean,",
          "$N$ for a population mean, $n-1$ for a variance. The <b>square root</b>, which",
          "is the only difference between $s^2$ and $s$. And whether the deviations are",
          "<b>squared</b> at all.",
          "<br><br>The two that get confused most often are the middle pair: $s^2$ and",
          "$s$ are the same expression apart from the root, so check for it every time.",
          "<br><br>Notice that the two <i>pairs</i> work the same way. A sample mean and a",
          "population mean are the same sum, divided by $n$ or by $N$; a sample proportion",
          "and a population proportion are the same count, divided by $n$ or by $N$. When",
          "the equation is handed to you without a label, the bottom of the fraction is",
          "what tells you whether you are looking at a statistic or a parameter."))
```

**(b)** *(notation)* Answer each in a few words.

```{r p03-notation}
show_tbl(data.frame(
  Question = c("What does $n$ stand for?",
               "What does $N$ stand for?",
               "What does the *hat* on $\\hat{p}$ tell you?",
               "What does $\\sum_{i=1}^{n} x_i$ mean?"),
  `Your answer` = cellv(c("the number of observations in the **sample**",
                          "the number of observations in the **population**",
                          "that it is computed from a sample, and estimates the parameter $p$",
                          "add up the observations, starting at the first and running through the $n^{th}$: $x_1 + x_2 + \\cdots + x_n$")),
  check.names = FALSE))
```

**(c)** Consider the following sample of five observations of a quantitative variable $X$:

```{r p03-data, results='asis'}
x3 <- c(2, 5, 6, 9, 13)
xbar3 <- mean(x3)
dev3  <- x3 - xbar3
ss3   <- sum(dev3^2)
var3  <- ss3 / (length(x3) - 1)
sd3   <- sqrt(var3)
stopifnot(length(x3) == 5, sum(x3) == 35, xbar3 == 7,
          abs(sum(dev3)) < 1e-12, ss3 == 70, var3 == 17.5,
          abs(sd3 - 4.1833) < 1e-4)
show_set(x3)
```

**Show how to compute** the mean $\bar{x}$ and the standard deviation $s$.

```{r p03-c, results='asis'}
why(paste("$\\bar{x} = \\dfrac{1}{n}\\sum_{i=1}^{n} x_i",
          "= \\dfrac{2 + 5 + 6 + 9 + 13}{5} = \\dfrac{35}{5} = 7$",
          "<br><br>$s = \\sqrt{\\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}}",
          "= \\sqrt{\\dfrac{(2-7)^2 + (5-7)^2 + \\cdots + (13-7)^2}{5-1}}",
          "= \\sqrt{\\dfrac{70}{4}} = \\sqrt{17.5} \\approx 4.18$"))
workspace(6)
```

**(d)** Now consider the following sample of ten observations of a qualitative variable
$Y$, the way each of ten students travelled to campus:

```{r p03-qual, results='asis'}
y3 <- c("Bus", "Walk", "Bike", "Bus", "Walk", "Bus", "Car", "Bus", "Walk", "Bike")
phat3 <- sum(y3 == "Bus") / length(y3)
stopifnot(length(y3) == 10, sum(y3 == "Bus") == 4, phat3 == 0.4)
show_set(y3, "Y")
```

**Show how to compute** the proportion of observations in the category **Bus**.

```{r p03-d, results='asis'}
why(paste("$\\hat{p} = \\dfrac{\\text{number of observations in the category}}{n}",
          "= \\dfrac{4}{10} = 0.4$",
          "<br><br>So 40% of the sample travelled by bus."))
workspace(4)
```

---

## Problem 4 — The distribution of a qualitative variable

A campus coffee stand recorded the drink ordered by each of 20 customers one morning:

```{r p04-data, results='asis'}
## Listed in the order they were served, NOT grouped by type -- students should
## have to tally them. Hard-coded rather than shuffled so the page renders the
## same list every time.
drinks <- c("Latte", "Drip", "Mocha", "Latte", "Drip", "Espresso", "Latte",
            "Drip", "Latte", "Mocha", "Drip", "Latte", "Espresso", "Drip",
            "Mocha", "Latte", "Drip", "Latte", "Mocha", "Latte")
d_tab <- table(factor(drinks, levels = c("Drip", "Espresso", "Latte", "Mocha")))
d_f   <- as.integer(d_tab)
d_rf  <- d_f / sum(d_f)
stopifnot(length(drinks) == 20, sum(d_f) == 20,
          identical(d_f, c(6L, 2L, 8L, 4L)),
          names(d_tab)[which.max(d_f)] == "Latte",
          abs(d_rf[3] - 0.40) < 1e-12)
show_set(drinks)
```

**(a)** Complete the frequency and relative frequency table.

```{r p04-table}
show_tbl(data.frame(
  Drink = names(d_tab),
  `Frequency $F(x)$` = cellv(d_f),
  `Relative frequency $RF(x)$` = cellv(sprintf("%.2f", d_rf)),
  check.names = FALSE))
```

**(b)** What is the modal category?

`r ans("Latte, with a frequency of 8")`

**(c)** Draw a bar graph of the distribution on the axes below.

```{r p04-bar, fig.height=3.6, fig.width=5.5}
bar_df <- data.frame(Drink = names(d_tab), Frequency = d_f)
p <- ggplot(bar_df, aes(x = Drink, y = Frequency)) +
  scale_y_continuous(limits = c(0, 9), breaks = 0:9) +
  theme_bw() + xlab("Drink") + ylab("Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p + geom_col(fill = "lightblue", colour = "black", width = 0.65)
} else {
  p + geom_blank()
}
```

**(d)** Would it make sense to add a *cumulative relative frequency* column to the table in
part (a)? Explain.

```{r p04-d, results='asis'}
why(paste("No. Cumulative relative frequency is the proportion of observations",
          "<i>less than or equal to</i> a given value, and that phrase only means",
          "something if the values can be put in order. Drink type is a <b>nominal</b>",
          "variable -- there is no sense in which a mocha is 'less than' a latte -- so",
          "there is no meaningful order to accumulate along.",
          "<br><br>Had the variable been ordinal, say drink size (small, medium, large),",
          "a cumulative relative frequency column would be perfectly sensible."))
workspace(3)
```

---

## Problem 5 — Frequency tables and the mean from a table

Twenty-five households were asked how many vehicles they own. The responses are summarized
below.

```{r p05-data}
v5 <- 0:4
f5 <- c(3, 9, 8, 4, 1)
rf5 <- f5 / sum(f5)
crf5 <- cumsum(rf5)
mean5 <- sum(v5 * f5) / sum(f5)
stopifnot(sum(f5) == 25, mean5 == 1.64,
          abs(crf5[3] - 0.80) < 1e-12,
          abs(sum(v5 * rf5) - mean5) < 1e-12)

## Which entries are left blank. The table is deliberately solvable: the missing
## frequency comes from the fact that the frequencies sum to n = 25 (or from
## CRF(3) - CRF(1) - RF(2), which is a second, independent route to it), and the
## missing CRF at x = 2 has to be filled in before part (e) can be read off it.
f_blank5   <- 4       # x = 3
rf_blank5  <- c(2, 4, 5)   # x = 1, 3, 4
crf_blank5 <- c(3, 5)      # x = 2, 4
stopifnot(sum(f5[-f_blank5]) == 21, f5[f_blank5] == 25 - 21,
          length(rf_blank5) == 3, length(crf_blank5) == 2)

show_tbl(data.frame(
  `Vehicles $x$` = v5,
  `Frequency $F(x)$` = cellsome(f5, f_blank5),
  `Relative frequency $RF(x)$` = cellsome(sprintf("%.2f", rf5), rf_blank5),
  `Cumulative relative frequency $CRF(x)$` = cellsome(sprintf("%.2f", crf5), crf_blank5),
  check.names = FALSE))
```

**(a)** Fill in the blank entries in the table. Remember that the frequencies must add to
$n$, the relative frequencies must add to $1$, and each cumulative relative frequency is
the running total of the relative frequencies above it.

```{r p05-a, results='asis'}
why(paste("<b>The missing frequency.</b> The frequencies have to add to $n = 25$, and the",
          "four that are given add to $3 + 9 + 8 + 1 = 21$, so $F(3) = 25 - 21 = 4$.",
          "<br><br><b>The missing relative frequencies.</b> $RF(x) = F(x)/n$:",
          "$RF(1) = 9/25 = 0.36$, $RF(3) = 4/25 = 0.16$, $RF(4) = 1/25 = 0.04$.",
          "As a check, the column now adds to",
          "$0.12 + 0.36 + 0.32 + 0.16 + 0.04 = 1.00$.",
          "<br><br><b>The missing cumulative relative frequencies.</b> Each one is the",
          "previous $CRF$ plus the current $RF$: $CRF(2) = 0.48 + 0.32 = 0.80$ and",
          "$CRF(4) = 0.96 + 0.04 = 1.00$. The last $CRF$ is always $1.00$ -- if yours",
          "is not, something above it is wrong."))
workspace(3)
```

**(b)** Show how to compute the mean number of vehicles using the **frequency** from the table
above.

```{r p05-b, results='asis'}
why(paste("$\\bar{x} = \\dfrac{1}{n}\\sum_{x} x\\,F(x)",
          "= \\dfrac{0(3) + 1(9) + 2(8) + 3(4) + 4(1)}{25}",
          "= \\dfrac{0 + 9 + 16 + 12 + 4}{25} = \\dfrac{41}{25} = 1.64$ vehicles."))
workspace(3)
```

**(c)** Now show how to compute the mean again using the **relative frequency** from the table
above, and confirm you get the same answer.

```{r p05-c, results='asis'}
why(paste("$\\bar{x} = \\sum_{x} x\\,RF(x)",
          "= 0(0.12) + 1(0.36) + 2(0.32) + 3(0.16) + 4(0.04)",
          "= 0 + 0.36 + 0.64 + 0.48 + 0.16 = 1.64$ vehicles.",
          "<br><br>The two formulas are the same calculation. Dividing by $n$ at the end",
          "(the frequency version) and dividing by $n$ at the start to build the relative",
          "frequencies (this version) get you to the same place. This is also why you",
          "must not divide by $n$ <i>again</i> in the relative frequency version."))
workspace(3)
```

**(d)** What proportion of these households own 2 or fewer vehicles?

`r ans("$0.80$")`

**(e)** Use the table to find the median number of vehicles owned.

```{r p05-e, results='asis'}
why(paste("<b>Median is 2 vehicles.</b> There are two ways to get there.",
          "<br><br><b>From the frequencies.</b> With $n = 25$ the median is the",
          "$13^{th}$ value once the data is put in order. The first $3$ values are $0$s",
          "and the next $9$ are $1$s, which takes you through the $12^{th}$ value, so the",
          "$13^{th}$ value is a $2$.",
          "<br><br><b>From the $CRF$ column.</b> The median is the $50^{th}$ percentile,",
          "so read down the $CRF$ column for where it first reaches $0.50$: it is $0.48$",
          "at $x = 1$ and $0.80$ at $x = 2$, so the median is $2$."))
workspace(3)
```

---

## Problem 6 — Binned frequency tables and histograms

The following are the times, in seconds, that 16 users took to complete a task on a website.

```{r p06-data, results='asis'}
t6 <- c(12.4, 14.1, 15.8, 16.2, 17.5, 18.0, 18.3, 19.1,
        20.2, 20.6, 21.7, 22.4, 23.9, 25.1, 26.8, 28.4)
brk6 <- c(12, 16, 20, 24, 28.5)
f6 <- as.integer(table(cut(t6, breaks = brk6, right = TRUE)))
stopifnot(length(t6) == 16, sum(f6) == 16,
          identical(f6, c(3L, 5L, 5L, 3L)))
show_set(t6)
```

**(a)** Complete the frequency table using the intervals given.

```{r p06-table}
show_tbl(data.frame(
  `Time (seconds)` = c("$12 < x \\leq 16$", "$16 < x \\leq 20$",
                       "$20 < x \\leq 24$", "$24 < x \\leq 28.5$"),
  `Frequency $F(x)$` = cellv(f6),
  `Relative frequency $RF(x)$` = cellv(sprintf("%.4f", f6 / sum(f6))),
  check.names = FALSE))
```

**(b)** Draw the histogram on the axes below.

```{r p06-hist, fig.height=3.8, fig.width=5.5}
hist_df <- data.frame(mid = c(14, 18, 22, 26.25), f = f6,
                      wid = diff(brk6))
p <- ggplot(hist_df, aes(x = mid, y = f, width = wid)) +
  scale_x_continuous(limits = c(12, 28.5), breaks = c(12, 16, 20, 24, 28.5)) +
  scale_y_continuous(limits = c(0, 6), breaks = 0:6) +
  theme_bw() + xlab("Time to complete task (seconds)") + ylab("Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p + geom_col(fill = "lightblue", colour = "black")
} else {
  p + geom_blank()
}
```

**(c)** Describe the shape of this distribution.

```{r p06-c, results="asis"}
why(paste("Roughly <b>symmetric and unimodal</b>, with the bulk of the observations in",
          "the two middle bins and the counts tapering off evenly at both ends. There is",
          "no long tail in either direction and no obvious gap, so nothing here looks",
          "like an outlier."))
workspace(3)
```

**(d)** A frequency table reports only how many observations fall in each interval.

  (i) Using **only** the table from part (a), what proportion of users finished the task in
  20 seconds or less?

  (ii) Name one quantity you could compute from the raw times listed above but **not** from
  the table alone, and say why the table cannot give it to you.

```{r p06-d, results="asis"}
why(paste("<b>(i)</b> The first two intervals cover everything up to 20 seconds, and they",
          "hold $3 + 5 = 8$ of the 16 users, so the proportion is $8/16 = 0.5$.",
          "<br><br><b>(ii)</b> The fastest or slowest individual time, or the exact mean or",
          "median. The table keeps only the <i>counts</i> per interval -- it says five users",
          "finished between 16 and 20 seconds, but not what any one of those five times was,",
          "so there is nothing left to take a maximum or an average of.",
          "<br><br>That loss of precision is the cost of binning. It buys a readable",
          "summary of the shape, which the raw list of 16 numbers does not give you."))
workspace(3)
```

---

## Problem 7 — Dot plots and stem-and-leaf plots

A biologist recorded the number of eggs in each of 15 birds' nests:

```{r p07-data, results='asis'}
e7 <- c(21, 24, 24, 27, 28, 31, 33, 33, 33, 35, 38, 41, 42, 44, 47)
mode7 <- as.integer(names(sort(table(e7), decreasing = TRUE))[1])
stopifnot(length(e7) == 15, mode7 == 33, identical(e7, sort(e7)))
show_set(e7)
```

**(a)** Construct a **stem-and-leaf plot**, using the tens digit as the stem.

```{r p07-stem, results='asis'}
if (SOLUTIONS) {
  stems <- 2:4
  rows <- vapply(stems, function(s) {
    leaves <- e7[e7 %/% 10 == s] %% 10
    sprintf("  %d | %s", s, paste(leaves, collapse = " "))
  }, character(1))
  cat('<pre style="color: red;">', paste(rows, collapse = "\n"), "</pre>", sep = "\n")
} else {
  cat('<pre>  2 |\n  3 |\n  4 |\n</pre>')
}
```

**(b)** Construct a **dot plot** on the number line below.

```{r p07-dot, fig.height=2.6, fig.width=6.5}
dot_df <- data.frame(x = e7)
dot_df$y <- ave(dot_df$x, dot_df$x, FUN = seq_along)
p <- ggplot(dot_df, aes(x = x, y = y)) +
  scale_x_continuous(limits = c(20, 48), breaks = seq(20, 48, 2)) +
  scale_y_continuous(limits = c(0.5, 4)) +
  theme_classic() + xlab("Eggs per nest") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), axis.line.y = element_blank())
if (SOLUTIONS) p + geom_point(size = 2.6) else p + geom_blank()
```

**(c)** What is the mode of this variable?

`r ans("33 eggs, which occurs three times")`

**(d)** Is the mode a measure of the center of this distribution? Explain.

```{r p07-d, results='asis'}
why(paste("No -- the mode is a measure of <b>location</b>. It tells you where the peak",
          "is, i.e. the most frequent value, not where the center is. Here the",
          "distribution happens to be roughly symmetric so the mode of 33 does land near",
          "the middle, but that is a coincidence of this data rather than a property of",
          "the mode.",
          "<br><br>For a skewed distribution the mode can sit well away from the center,",
          "and a bimodal distribution has two modes with the center usually falling in",
          "the gap between them, where the data are sparsest."))
workspace(3)
```

---

## Problem 8 — Center, and what an outlier does to it

Consider the following five observations of a variable $X$, the number of minutes five
customers spent in a shop:

```{r p08-data, results='asis'}
x8 <- c(6, 6, 9, 12, 17)
mean8 <- mean(x8); med8 <- median(x8)
mode8 <- as.integer(names(sort(table(x8), decreasing = TRUE))[1])
x8b <- c(x8, 70)
mean8b <- mean(x8b); med8b <- median(x8b)
stopifnot(length(x8) == 5, sum(x8) == 50, mean8 == 10, med8 == 9, mode8 == 6,
          length(x8b) == 6, sum(x8b) == 120, mean8b == 20, med8b == 10.5,
          mean8b > max(x8))
show_set(x8)
```

**(a)** Show how to compute the mean, the median and the mode.

```{r p08-a}
show_tbl(data.frame(
  Statistic = c("Mean $\\bar{x}$", "Median", "Mode"),
  `Your answer` = cellbox(c(
    "$\\bar{x} = \\frac{1}{n}\\sum_{i=1}^{n} x_i = \\frac{6 + 6 + 9 + 12 + 17}{5} = \\frac{50}{5} = 10$ minutes",
    "the $3^{rd}$ of 5 ordered values: 9 minutes",
    "the most frequent value: 6 minutes")),
  check.names = FALSE)) %>%
  column_spec(2, width = "4.5in")
```

**(b)** A sixth customer stays for **70 minutes**. Recompute the mean and the median with that
observation included. Which of the two moved more, and why?

```{r p08-b, results='asis'}
why(paste("$\\bar{x} = \\frac{1}{n}\\sum_{i=1}^{n} x_i",
          "= \\frac{6 + 6 + 9 + 12 + 17 + 70}{6} = \\frac{120}{6} = 20$ minutes.",
          "<br><br>With $n = 6$ there is no single middle value, so the median is the",
          "average of the $3^{rd}$ and $4^{th}$ ordered values:",
          "$\\frac{9 + 12}{2} = 10.5$ minutes.",
          "<br><br>The <b>mean</b> moved far more -- it doubled, from 10 up to 20 minutes",
          "-- while the median moved from 9 to 10.5, a shift of a minute and a half.",
          "<br><br>The mean uses the <i>value</i> of every observation, so a single very",
          "large value pulls it upward in proportion to how extreme it is. The median",
          "depends only on <i>position</i> in the ordered data; adding one more value at",
          "the top shifts the middle position over by half a place and nothing more.",
          "<br><br>Notice too that after the outlier the mean of 20 minutes is larger than",
          "<i>every observation except the outlier itself</i>. It is no longer describing",
          "a typical customer at all."))
workspace(5)
```

**(c)** Which of the two is **resistant**? Give a one-sentence definition of that word.

`r ans("The median is resistant.")`

```{r p08-c, results='asis'}
why(paste("A summary measure is <b>resistant</b> when a single extreme observation barely",
          "moves it. Whether a measure is resistant comes down to one thing: whether it",
          "uses the values of every observation, or only positions in the ordered data."))
workspace(2)
```

---

## Problem 9 — Percentiles and quartiles

The following are the ages of the 10 members of a community board, sorted:

```{r p09-data, results='asis'}
a9 <- c(29, 34, 38, 41, 45, 49, 52, 56, 61, 67)
q9 <- q_halves(a9)
iqr9 <- q9[["Q3"]] - q9[["Q1"]]
## Part (c) adds an 11th member, so the same data is worked once with n even (no
## median to drop) and once with n odd (drop the median) -- the two halves of the
## rule, on one dataset.
a9b <- sort(c(a9, 58))
q9b <- q_halves(a9b)
stopifnot(length(a9) == 10, identical(a9, sort(a9)),
          q9[["Q1"]] == 38, q9[["Q2"]] == 47, q9[["Q3"]] == 56, iqr9 == 18,
          length(a9b) == 11,
          q9b[["Q1"]] == 38, q9b[["Q2"]] == 49, q9b[["Q3"]] == 58,
          sum(a9 <= 61) == 9)
show_set(a9)
```

**(a)** Find the median, $Q2$.

`r ans("$Q2 = (45 + 49)/2 = 47$, the average of the $5^{th}$ and $6^{th}$ of 10 ordered values")`

**(b)** Find $Q1$ and $Q3$ using the **method of halves** we use in this class.

```{r p09-b, results='asis'}
why(paste("Because $n = 10$ is even there is no middle value to set aside -- the data",
          "splits cleanly into two halves of five. The lower half is",
          "$\\{29, 34, 38, 41, 45\\}$ and the upper half is $\\{49, 52, 56, 61, 67\\}$, so",
          "each quartile is just the middle value of its half:",
          "<br><br>$Q1 = 38$ years and $Q3 = 56$ years."))
workspace(4)
```

**(c)** An eleventh member, aged $58$, joins the board. Find $Q1$, $Q2$ and $Q3$ for the
new set of 11 ages. Remember that when $n$ is odd the median is **left out** of both halves.

```{r p09-c, results='asis'}
why(paste("The ordered ages are now",
          "$\\{29, 34, 38, 41, 45, 49, 52, 56, 58, 61, 67\\}$.",
          "<br><br>With $n = 11$ the median is the single middle value, the $6^{th}$ age:",
          "$Q2 = 49$ years. Because $n$ is odd, that value is now dropped from both halves.",
          "<br><br>The lower half is $\\{29, 34, 38, 41, 45\\}$, so $Q1 = 38$ years.",
          "<br>The upper half is $\\{52, 56, 58, 61, 67\\}$, so $Q3 = 58$ years.",
          "<br><br>Compare this with part (b): the method is the same both times. The",
          "only thing that changes is that an odd $n$ leaves you a middle value to drop,",
          "and an even $n$ does not."))
workspace(5)
```

**(d)** Using the original 10 ages, what is the $IQR$, and what does it measure?

`r ans("$IQR = Q3 - Q1 = 56 - 38 = 18$ years")`

```{r p09-d, results='asis'}
why(paste("The IQR is the distance between $Q1$ and $Q3$, so it describes how spread out",
          "the <b>middle 50%</b> of the data are. Because it ignores the extreme ends of",
          "the distribution entirely, it is highly resistant to outliers."))
workspace(2)
```

**(e)** In the original 10 ages, the board member aged 61 is at roughly what percentile?
Answer in a sentence that says what that means.

```{r p09-e, results='asis'}
why(paste("Nine of the 10 members are aged 61 or younger, so 61 sits at about the",
          "$9/10 = 90^{th}$ percentile: roughly 90% of the board is at or below that age,",
          "and only one member is older."))
workspace(3)
```

---

## Problem 10 — Deviations, variance and standard deviation

A quality inspector took three packets of ground coffee off a filling machine and weighed
each one, in grams:

```{r p10-data, results='asis'}
w10 <- c(48, 54, 57)
m10 <- mean(w10)
dev10 <- w10 - m10
ss10 <- sum(dev10^2)
var10 <- ss10 / (length(w10) - 1)
sd10 <- sqrt(var10)
new10 <- 63                      # the fourth packet, in part (b)
stopifnot(length(w10) == 3, sum(w10) == 159, m10 == 53,
          abs(sum(dev10)) < 1e-12, ss10 == 42, var10 == 21,
          abs(sd10 - 4.5826) < 1e-3,
          new10 - m10 == 10, diff(range(w10)) == 9)
show_set(w10)
```

The sample mean of these three packets is $\bar{x} = \dfrac{159}{3} = 53$ grams.

**(a)** Complete the deviation table below, then add up the deviation column.

```{r p10-table}
show_tbl(data.frame(
  `$x_i$` = w10,
  `$x_i - \\bar{x}$` = cellbox(sprintf("%.0f", dev10), height = "0.7in"),
  `$(x_i - \\bar{x})^2$` = cellbox(sprintf("%.0f", dev10^2), height = "0.7in"),
  check.names = FALSE)) %>%
  column_spec(1:3, width = "1.6in")
```

```{r p10-a, results='asis'}
why(paste("The deviations are $48 - 53 = -5$, $54 - 53 = 1$ and $57 - 53 = 4$ grams, and",
          "their squares are $25$, $1$ and $16$.",
          "<br><br>The deviation column sums to <b>0</b>, and the squared deviation column",
          "sums to $25 + 1 + 16 = 42$.",
          "<br><br>The deviations summing to zero is not a property of these three numbers",
          "-- it happens for every dataset, because the mean is the balancing point. That",
          "is exactly why the variance squares the deviations before adding them: without",
          "squaring, the positives and negatives always cancel and you learn nothing."))
workspace(3)
```

**(b)** A fourth packet comes off the machine weighing **63 grams**. Using the sample mean
$\bar{x} = 53$ grams from the three packets above, compute this packet's deviation from the
mean and interpret what that number tells you.

```{r p10-b, results='asis'}
why(paste("$x - \\bar{x} = 63 - 53 = +10$ grams.",
          "<br><br>This packet weighs 10 grams <i>more</i> than the average of the three",
          "packets already measured. A deviation carries both a size and a sign: the size",
          "is how far the observation sits from the mean in the units of the data (grams",
          "here), and the positive sign says it sits <i>above</i> the mean. A packet",
          "weighing 43 grams would have deviation $-10$ grams -- the same distance from",
          "the mean, on the other side."))
workspace(3)
```

**(c)** Show or explain how to compute the sample variance $s^2$ from the table you filled in
above, and then the sample standard deviation $s$.

```{r p10-c, results='asis'}
why(paste("The squared deviation column is the sum the variance formula needs. Add that",
          "column up and divide by $n - 1$:",
          "<br><br>$s^{2} = \\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}",
          "= \\dfrac{25 + 1 + 16}{3 - 1} = \\dfrac{42}{2} = 21$ grams squared",
          "<br><br>$s = \\sqrt{s^{2}} = \\sqrt{21} \\approx 4.58$ grams"))
workspace(3)
```

**(d)** State the units of $\bar{x}$, of $s^2$ and of $s$.

```{r p10-d}
show_tbl(data.frame(
  Quantity = c("$\\bar{x}$", "$s^2$", "$s$"),
  Units = cellv(c("grams", "grams squared", "grams")),
  check.names = FALSE))
```

```{r p10-d2, results='asis'}
why(paste("Squaring the deviations squares their units too, which is why the variance is",
          "in grams squared -- a quantity nobody can picture. Taking the square root",
          "undoes it and puts $s$ back on the scale of the original data, where it reads",
          "as roughly the typical distance of a packet from the mean weight."))
workspace(2)
```

**(e)** Show how to compute the range. Which is the better description of the spread of this data, the
range or the standard deviation, and why?

```{r p10-e, results='asis'}
why(paste("Range $= \\max x - \\min x = 57 - 48 = 9$ grams.",
          "<br><br>The <b>standard deviation</b> is the better description here. The range",
          "is built from the two most extreme observations and <i>nothing else</i>: here it",
          "ignores the middle packet entirely, and in a larger sample it would ignore almost",
          "all of the data. It also moves by the full amount if either endpoint changes.",
          "The standard deviation uses every observation.",
          "<br><br>The range is worth reporting as a quick descriptive fact, but it is",
          "almost never used as <i>the</i> measure of spread."))
workspace(3)
```

---

## Problem 11 — Five-number summary, boxplots and outliers

A clinic recorded the number of days between a referral and an appointment for 12 patients:

```{r p11-data, results='asis'}
d11 <- c(3, 5, 6, 8, 9, 11, 12, 14, 15, 17, 19, 58)
q11 <- q_halves(d11)
iqr11 <- q11[["Q3"]] - q11[["Q1"]]
lo11 <- q11[["Q1"]] - 1.5 * iqr11
hi11 <- q11[["Q3"]] + 1.5 * iqr11
out11 <- d11[d11 < lo11 | d11 > hi11]
whisk_hi <- max(d11[d11 <= hi11])

## n is EVEN on purpose. base boxplot() draws Tukey's hinges, which agree with
## the class's median-of-halves rule only when n is even -- with an odd n the
## drawn box would contradict the five-number summary in part (a). Asserted
## rather than trusted.
stopifnot(length(d11) == 12, identical(d11, sort(d11)),
          all.equal(as.numeric(fivenum(d11)[c(2, 4)]),
                    c(q11[["Q1"]], q11[["Q3"]])),
          q11[["Q1"]] == 7, q11[["Q2"]] == 11.5, q11[["Q3"]] == 16,
          iqr11 == 9, lo11 == -6.5, hi11 == 29.5,
          identical(out11, 58), whisk_hi == 19,
          abs(mean(d11) - 14.75) < 1e-9)
show_set(d11)
```

**(a)** Give the five-number summary. Use the method of halves for the quartiles.

```{r p11-a}
show_tbl(data.frame(
  Minimum = cellv("3"), Q1 = cellv("7"), Median = cellv("11.5"),
  Q3 = cellv("16"), Maximum = cellv("58"), check.names = FALSE))
```

```{r p11-a2, results='asis'}
why(paste("$n = 12$ is even, so there is no middle value to leave out: the median is the",
          "average of the $6^{th}$ and $7^{th}$ values, $Q2 = (11+12)/2 = 11.5$ days, and",
          "the data splits cleanly into two halves of six.",
          "<br><br>The lower half is $\\{3,5,6,8,9,11\\}$ with median",
          "$Q1 = (6+8)/2 = 7$; the upper half is $\\{12,14,15,17,19,58\\}$ with median",
          "$Q3 = (15+17)/2 = 16$."))
workspace(3)
```

**(b)** Show how to compute the $IQR$ and both outlier boundaries.

```{r p11-b, results='asis'}
why(paste("$IQR = Q3 - Q1 = 16 - 7 = 9$ days, so $1.5 \\times IQR = 13.5$.",
          "<br><br>Lower boundary $= Q1 - 1.5 \\times IQR = 7 - 13.5 = -6.5$ days.",
          "<br>Upper boundary $= Q3 + 1.5 \\times IQR = 16 + 13.5 = 29.5$ days.",
          "<br><br>Nothing can fall below $-6.5$ days, so only the upper boundary can",
          "flag anything here."))
workspace(3)
```

**(c)** Which observations, if any, does the $1.5 \times IQR$ rule flag as outliers?

`r ans("Just one: the 58-day wait, which is above the upper boundary of 29.5")`

**(d)** Draw the boxplot on the axis below. Be careful about where the right whisker ends.

```{r p11-box, fig.height=2.8, fig.width=6.5}
if (SOLUTIONS) {
  par(mar = c(4, 1, 1, 1))
  boxplot(d11, horizontal = TRUE, axes = FALSE, staplewex = 1, col = "lightgrey",
          ylim = c(0, 60))
  axis(1, at = seq(0, 60, 5))
  mtext("Days from referral to appointment", side = 1, line = 2.5)
} else {
  par(mar = c(4, 1, 1, 1))
  plot(NA, xlim = c(0, 60), ylim = c(0.5, 1.5), axes = FALSE, xlab = "", ylab = "")
  axis(1, at = seq(0, 60, 5))
  mtext("Days from referral to appointment", side = 1, line = 2.5)
}
```

```{r p11-d, results='asis'}
why(paste("The right whisker stops at <b>19</b>, the largest observation that is still at",
          "or below the upper boundary of 29.5 -- <i>not</i> at the maximum of 58, and",
          "not at the boundary itself. A whisker always ends on an actual data value. The",
          "58 is drawn as its own point beyond the whisker.",
          "<br><br>The left whisker runs all the way out to the minimum of 3, because",
          "nothing is flagged at the low end."))
workspace(3)
```

**(e)** Describe the shape of this distribution from the boxplot, and say what that implies
about the mean relative to the median.

```{r p11-e, results='asis'}
why(paste("The distribution is <b>skewed right</b>: the median sits left of centre in the",
          "box, the right whisker is longer than the left, and there is a high outlier.",
          "<br><br>A right skew pulls the mean toward the long tail, so the",
          "<b>mean will be greater than the median</b>. The mean here is 14.75 days",
          "against a median of 11.5."))
workspace(3)
```

**(f)** The clinic manager wants to delete the 58-day wait because it is "clearly a
mistake." What would you tell her?

```{r p11-f, results='asis'}
why(paste("That being flagged by the $1.5 \\times IQR$ rule is an <b>arithmetic result,",
          "not a verdict</b>. The rule reports only that a value sits unusually far from",
          "the middle half of the data; it cannot distinguish a recording error from a",
          "genuinely unusual case, and it marks the observation as worth a second look",
          "rather than as wrong.",
          "<br><br>A 58-day wait is entirely plausible -- it may be the single most",
          "important number in the dataset for a manager trying to improve scheduling.",
          "Deleting it because a rule flagged it would throw away real information."))
workspace(4)
```

---

## Problem 12 — Cumulative distributions

A librarian recorded the number of books checked out by each of 20 patrons in one hour.

```{r p12-data}
x12 <- 0:5
## Deliberately, visibly right skewed: a tall pile at 0 and 1 and a thin tail out
## to 5. Part (e) asks for the shape and for mean vs median, so the skew has to be
## strong enough that "roughly symmetric" is not an equally good answer -- here the
## mean sits 0.4 books above the median, which is plain both in the frequencies
## and in how quickly the step plot flattens out after x = 1.
f12 <- c(6, 7, 3, 2, 1, 1)
rf12 <- f12 / sum(f12)
crf12 <- cumsum(rf12)
tab12 <- data.frame(x = x12, crf = crf12)
raw12 <- rep(x12, f12)
q12 <- q_halves(raw12)
## No CRF may land exactly on 0.25, 0.50 or 0.75: part (c) is read off the plot,
## and a step that ends ON the line has no unambiguous reading.
stopifnot(sum(f12) == 20, length(raw12) == 20,
          abs(crf12[3] - 0.80) < 1e-12,
          !any(abs(outer(crf12, c(0.25, 0.5, 0.75), "-")) < 1e-12),
          q12[["Q1"]] == 0, q12[["Q2"]] == 1, q12[["Q3"]] == 2,
          sum(x12 * f12) == 28, mean(raw12) == 1.4,
          mean(raw12) > q12[["Q2"]])
```

**(a)** Complete the table.

```{r p12-table}
show_tbl(data.frame(
  `Books $x$` = x12,
  `Frequency $F(x)$` = f12,
  `Relative frequency $RF(x)$` = cellv(sprintf("%.2f", rf12)),
  `Cumulative relative frequency $CRF(x)$` = cellv(sprintf("%.2f", crf12)),
  check.names = FALSE))
```

**(b)** Plot the cumulative distribution on the axes below. Remember it is a **step
function**: flat between observed values, jumping at each one.

```{r p12-cumdist, fig.height=4.6, fig.width=7}
lead12 <- data.frame(x = -1, crf = 0)
p <- ggplot() +
  scale_x_continuous(limits = c(-1, 6), breaks = 0:6) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.05)) +
  theme_bw() + xlab("Books checked out") + ylab("Cumulative Relative Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p +
    geom_step(data = rbind(lead12, tab12), aes(x = x, y = crf),
              direction = "hv", linewidth = 1) +
    geom_point(data = tab12, aes(x = x, y = crf), size = 3)
} else {
  p + geom_blank()
}
```

**(c)** Use the plot to find $Q1$, the median, and $Q3$.

```{r p12-c, results='asis'}
why(paste("Read across from $0.25$, $0.50$ and $0.75$ on the $y$-axis to the curve, then",
          "drop to the $x$-axis. Each of the three readings crosses a jump:",
          "<br><br>$Q1 = 0$ books, $Q2 = 1$ book, $Q3 = 2$ books.",
          "<br><br>Check them against halving the ordered data -- they agree. The",
          "cumulative distribution is not a new definition of a quartile, just a faster",
          "way to read one off, and unlike halving it gives you <i>any</i> percentile."))
workspace(4)
```

**(d)** What proportion of patrons checked out 2 or fewer books?

`r ans("$0.80$, read straight off the $CRF$ column at $x = 2$")`

**(e)** Describe the shape of this distribution, and say whether the mean will be greater
than, less than, or equal to the median.

```{r p12-e, results='asis'}
why(paste("The distribution is <b>skewed right</b>: most of the patrons are piled up at",
          "0 and 1 book -- $13$ of the $20$, so nearly two thirds of them -- and the",
          "frequencies then fall away to a thin tail of $3, 2, 1, 1$ stretching out to 5",
          "books. On the cumulative plot the same thing shows up as a curve that climbs",
          "steeply over the first two jumps and is almost flat after that.",
          "<br><br>A right skew pulls the mean toward the long tail while leaving the",
          "median where it is, so the <b>mean is greater than the median</b>. Here the",
          "mean is $28/20 = 1.4$ books against a median of 1 -- the handful of patrons",
          "checking out 4 and 5 books drags the mean up, but they move the middle value",
          "not at all."))
workspace(3)
```

---

```{r review-outro, results='asis'}
if (!SOLUTIONS) {
  cat(paste(
    "## Check your work\n\n",
    "Worked solutions are posted under",
    "**Exam Study Guides &rarr; Exam 1 Review &rarr; Solutions**. Work each problem",
    "through before looking at them.\n\n",
    "Every quartile on this exam uses the method of halves. Problem 9 works it both",
    "ways: with an odd $n$, where the median is dropped, and with an even $n$, where",
    "it is not.\n"))
} else {
  cat(paste(
    "## A note on the exam\n\n",
    "These solutions do more arithmetic than the exam requires. Exam 1 asks you to",
    "interpret values, identify equations and read plots. The formulas are shown during",
    "the exam, but **unlabeled** -- identifying which is which is on you.\n"))
}
```
