<!--
  SHARED BODY for the Exam 1 review problems.

  This file is NOT a page. It is included as a child document by both
  exam1_review_problems.Rmd and exam1_review_problems_solns.Rmd, each of which
  sets the SOLUTIONS flag before including it. Writing the questions once means
  an answer can never drift away from the question it belongs to -- the two
  pages were previously ~300 lines of hand-synced near-duplicate.

  The leading underscore keeps docs/build_site.R from rendering it as a page of
  its own (see the `^[^_]` pattern in its subdirectory loop).

  To add a problem: write the question once, and wrap every answer in ans() for
  short fill-ins, or an `eval = SOLUTIONS` chunk for worked steps. Compute the
  answer from the same vector the question prints, and stopifnot() anything you
  also state in prose.
-->

```{r review-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggthemes)
library(ggpubr)

stopifnot(exists("SOLUTIONS"), is.logical(SOLUTIONS), length(SOLUTIONS) == 1L)

## Short inline answer. Renders a blank in the problems copy and red text in the
## solutions copy -- the same red the Spring 2024 solutions used.
ans <- function(x, blank = "$\\_\\_\\_\\_\\_\\_\\_\\_$") {
  if (SOLUTIONS) sprintf('<span style="color: red;">%s</span>', x) else blank
}

## A whole answer paragraph, for parts that want a sentence rather than a value.
## Emits nothing at all in the problems copy.
why <- function(x) {
  if (SOLUTIONS) cat(sprintf('\n\n<span style="color: red;">%s</span>\n\n', x))
}

## Vertical space for students to work in, on the problems copy only.
workspace <- function(lines = 4) {
  if (!SOLUTIONS) cat(strrep("<br>\n", lines))
}

## Standard table look, matching the rest of the site.
show_tbl <- function(d, ...) {
  kable(d, format = "html", row.names = FALSE, escape = FALSE, ...) %>%
    kable_classic(full_width = FALSE, html_font = "Cambria") %>%
    kable_styling(bootstrap_options = "striped")
}

## A cell that is blank for students and filled in red for the key. Vectorised so
## a whole column can be built at once.
cellv <- function(x) {
  if (SOLUTIONS) sprintf('<span style="color: red;">%s</span>', x) else ""
}

## Print a set of observations in the notation used in the lecture notes:
##   X = {2, 5, 6, 9, 13}
## Character values are wrapped in \text{} so they typeset upright rather than as
## italic maths variables. Every problem that hands students raw data uses this,
## so the review set reads the same way the notes do.
show_set <- function(v, name = "X") {
  vals <- if (is.character(v)) sprintf("\\text{%s}", v) else format(v, trim = TRUE)
  cat(sprintf("\n\n$$%s = \\{ %s \\}$$\n\n", name, paste(vals, collapse = ", \\ ")))
}

## The class's quartile rule: median of halves, dropping the median itself when n
## is odd. Used by every problem that needs a quartile, so no problem can quietly
## disagree with Q16 of the exam or with the Week 3 notes.
q_halves <- function(x) {
  x <- sort(x); n <- length(x); h <- n %/% 2
  c(Q1 = median(x[1:h]), Q2 = median(x), Q3 = median(x[(n - h + 1):n]))
}
```

```{r review-intro, results='asis'}
if (SOLUTIONS) {
  cat('Worked solutions. Answers are shown <span style="color: red;">in red</span>.\n\n')
}
```

These problems cover everything on Exam 1: Weeks 1 through 3, up to and including
cumulative distributions. They are **more computational than the exam** by design. The exam
asks you to interpret values, identify equations and read plots; these ask you to work the
arithmetic behind them.

Work them by hand. Every dataset is small enough to do on paper, and a calculator is fine.

---

## Problem 1 — The anatomy of data

A researcher recorded the following information for each of six trees in a city park.

```{r p01-data}
trees <- data.frame(
  `Tree ID`           = c("T-01", "T-02", "T-03", "T-04", "T-05", "T-06"),
  `Species`           = c("Maple", "Oak", "Maple", "Pine", "Oak", "Maple"),
  `Health rating`     = c("Good", "Fair", "Excellent", "Good", "Poor", "Good"),
  `Branches removed`  = c(3, 0, 5, 1, 8, 2),
  `Trunk width (cm)`  = sprintf("%.1f", c(31.4, 58.2, 22.7, 44.0, 61.9, 27.5)),
  check.names = FALSE, stringsAsFactors = FALSE
)
show_tbl(trees)
```

**(a)** How many observations are in this data table, and how many variables?

`r ans("6 observations and 5 variables")`

**(b)** What is the value of $n$ for this sample?

`r ans("$n = 6$")`

**(c)** Classify each of the four measured variables as nominal, ordinal, discrete or
continuous.

```{r p01-types}
show_tbl(data.frame(
  Variable = c("Species", "Health rating", "Branches removed", "Trunk width (cm)"),
  `Your answer` = cellv(c("nominal", "ordinal", "discrete", "continuous")),
  check.names = FALSE))
```

```{r p01-why, results='asis'}
why(paste("<b>Species</b> is a category with no natural ordering, so nominal.",
          "<b>Health rating</b> is a category <i>with</i> an ordering -- poor, fair, good,",
          "excellent -- so ordinal. <b>Branches removed</b> counts things, so it takes",
          "whole-number values only: discrete. <b>Trunk width</b> is a measurement on a",
          "scale, recorded to a tenth of a centimetre, so continuous.",
          "<br><br>Note that <i>Tree ID</i> is a label, not a measured variable -- it",
          "identifies the observation rather than describing it."))
```

**(d)** These six trees are a sample. Describe the population the researcher is presumably
interested in, and say whether $N$ is known.

```{r p01-pop, results='asis'}
why(paste("The population is all of the trees in the city park (or, more broadly, all",
          "trees the park service manages). $N$ is the number of trees in that",
          "population. Here it is not given -- and unlike most populations we will meet,",
          "this one is finite and could actually be counted."))
workspace(3)
```

---

## Problem 2 — Statistics, parameters, and inference

A city has 8,500 registered bicycles. A researcher takes a random sample of 200 of them and
finds that 46 have no working lights. She also measures the age of each sampled bicycle and
finds a mean age of 6.2 years.

```{r p02-calc}
N_bikes <- 8500; n_bikes <- 200; no_lights <- 46
phat <- no_lights / n_bikes
stopifnot(phat == 0.23)
```

**(a)** Give the values of $N$ and $n$.

`r ans("$N = 8{,}500$ and $n = 200$")`

**(b)** Compute the sample proportion of bicycles with no working lights. Which symbol
denotes it?

`r ans("$\\hat{p} = 46/200 = 0.23$")`

```{r p02-b, results='asis'}
why(paste("The <b>sample</b> proportion is written $\\hat{p}$. The hat marks it as a",
          "quantity computed from a sample. The corresponding population proportion --",
          "the fraction of all 8,500 bicycles with no working lights -- is written $p$",
          "and is unknown."))
workspace(2)
```

**(c)** For each quantity, say whether it is a **statistic** or a **parameter**, and give
the symbol that denotes it.

```{r p02-symbols}
show_tbl(data.frame(
  Quantity = c("The mean age of the 200 sampled bicycles",
               "The mean age of all 8,500 registered bicycles",
               "The proportion of the 200 sampled with no lights",
               "The proportion of all 8,500 with no lights"),
  `Statistic or parameter` = cellv(c("statistic", "parameter", "statistic", "parameter")),
  Symbol = cellv(c("$\\bar{x}$", "$\\mu$", "$\\hat{p}$", "$p$")),
  check.names = FALSE))
```

**(d)** The researcher reports the mean age of the 200 bicycles she measured, then uses it
to estimate the mean age of all 8,500. Which of those two acts is descriptive statistics
and which is inferential? Explain what makes the difference.

```{r p02-d, results='asis'}
why(paste("Reporting the mean of the 200 measured bicycles is <b>descriptive</b>: it",
          "summarizes the data in hand and claims nothing beyond it. Using that mean to",
          "estimate the mean age of all 8,500 is <b>inferential</b>: it reaches past the",
          "sample to a population that was never fully observed.",
          "<br><br>The difference is not which data you have -- both acts use the same",
          "200 bicycles. It is <i>what you are claiming</i>: whether the conclusion stops",
          "at the sample or extends beyond it."))
workspace(4)
```

---

## Problem 3 — Reading the formulas

On the exam the formulas are provided **unlabeled** — the equations only, with no names.
This problem is practice at telling them apart.

**(a)** Each equation below computes one of the quantities in the list. Write the name of
the quantity next to its equation.

> sample mean &nbsp;&middot;&nbsp; population mean &nbsp;&middot;&nbsp; sample proportion
> &nbsp;&middot;&nbsp; sample variance &nbsp;&middot;&nbsp; sample standard deviation
> &nbsp;&middot;&nbsp; interquartile range

```{r p03-match}
show_tbl(data.frame(
  Equation = c("$\\dfrac{1}{n}\\sum_{i=1}^{n} x_i$",
               "$\\sqrt{\\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}}$",
               "$Q3 - Q1$",
               "$\\dfrac{1}{N}\\sum_{i=1}^{N} x_i$",
               "$\\dfrac{\\text{number of observations in the category}}{n}$",
               "$\\dfrac{1}{n-1}\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}$"),
  `What it computes` = cellv(c("sample mean", "sample standard deviation",
                               "interquartile range", "population mean",
                               "sample proportion", "sample variance")),
  Symbol = cellv(c("$\\bar{x}$", "$s$", "$IQR$", "$\\mu$", "$\\hat{p}$", "$s^2$")),
  check.names = FALSE))
```

```{r p03-a, results='asis'}
why(paste("Three things give each one away. The <b>divisor</b>: $n$ for a sample mean,",
          "$N$ for a population mean, $n-1$ for a variance. The <b>square root</b>, which",
          "is the only difference between $s^2$ and $s$. And whether the deviations are",
          "<b>squared</b> at all.",
          "<br><br>The two that get confused most often are the middle pair: $s^2$ and",
          "$s$ are the same expression apart from the root, so check for it every time."))
```

**(b)** *(notation)* Answer each in a few words.

```{r p03-notation}
show_tbl(data.frame(
  Question = c("What does $n$ stand for?",
               "What does $N$ stand for?",
               "What does the *hat* on $\\hat{p}$ tell you?",
               "In $\\sum_{i=1}^{n} x_i$, what does the $i=1$ below the $\\Sigma$ mean?"),
  `Your answer` = cellv(c("the number of observations in the **sample**",
                          "the number of observations in the **population**",
                          "that it is computed from a sample, and estimates the parameter $p$",
                          "start the count at the first observation and add up to the $n^{th}$")),
  check.names = FALSE))
```

**(c)** Consider the following sample of five observations of a quantitative variable $X$:

```{r p03-data, results='asis'}
x3 <- c(2, 5, 6, 9, 13)
xbar3 <- mean(x3)
dev3  <- x3 - xbar3
ss3   <- sum(dev3^2)
var3  <- ss3 / (length(x3) - 1)
sd3   <- sqrt(var3)
stopifnot(length(x3) == 5, sum(x3) == 35, xbar3 == 7,
          abs(sum(dev3)) < 1e-12, ss3 == 70, var3 == 17.5,
          abs(sd3 - 4.1833) < 1e-4)
show_set(x3)
```

**Show how to compute** the mean $\bar{x}$ and the standard deviation $s$.

```{r p03-c, results='asis'}
why(paste("<b>Mean.</b>",
          "$\\bar{x} = \\frac{1}{n}\\sum_{i=1}^{n} x_i",
          "= \\frac{2 + 5 + 6 + 9 + 13}{5} = \\frac{35}{5} = 7$",
          "<br><br><b>Standard deviation.</b> Take the deviation of each observation from",
          "the mean, square it, and add:",
          "<br><br>$\\sum_{i=1}^{n}(x_i - \\bar{x})^{2}",
          "= (-5)^2 + (-2)^2 + (-1)^2 + 2^2 + 6^2",
          "= 25 + 4 + 1 + 4 + 36 = 70$",
          "<br><br>Divide by $n - 1$ to get the variance, then take the square root:",
          "<br><br>$s^{2} = \\frac{70}{5 - 1} = \\frac{70}{4} = 17.5",
          "\\qquad s = \\sqrt{17.5} \\approx 4.18$",
          "<br><br>Note the deviations themselves -- $-5, -2, -1, 2, 6$ -- add to $0$.",
          "That happens for <b>every</b> dataset, because the mean is the balancing",
          "point, and it is why the deviations must be squared before they are added."))
workspace(6)
```

**(d)** Now consider the following sample of ten observations of a qualitative variable
$Y$, the way each of ten students travelled to campus:

```{r p03-qual, results='asis'}
y3 <- c("Bus", "Walk", "Bike", "Bus", "Walk", "Bus", "Car", "Bus", "Walk", "Bike")
phat3 <- sum(y3 == "Bus") / length(y3)
stopifnot(length(y3) == 10, sum(y3 == "Bus") == 4, phat3 == 0.4)
show_set(y3, "Y")
```

**Show how to compute** the proportion of observations in the category **Bus**.

```{r p03-d, results='asis'}
why(paste("Count the observations in the category, then divide by the sample size:",
          "<br><br>$\\hat{p} = \\frac{\\text{number of observations in the category}}{n}",
          "= \\frac{4}{10} = 0.4$",
          "<br><br>So 40% of the sample travelled by bus. Note that a proportion is the",
          "one summary here that <i>can</i> be computed for a qualitative variable -- a",
          "mean and a standard deviation cannot, because the categories are not numbers."))
workspace(4)
```

---

## Problem 4 — The distribution of a qualitative variable

A campus coffee stand recorded the drink ordered by each of 20 customers one morning:

```{r p04-data, results='asis'}
## Listed in the order they were served, NOT grouped by type -- students should
## have to tally them. Hard-coded rather than shuffled so the page renders the
## same list every time.
drinks <- c("Latte", "Drip", "Mocha", "Latte", "Drip", "Espresso", "Latte",
            "Drip", "Latte", "Mocha", "Drip", "Latte", "Espresso", "Drip",
            "Mocha", "Latte", "Drip", "Latte", "Mocha", "Latte")
d_tab <- table(factor(drinks, levels = c("Drip", "Espresso", "Latte", "Mocha")))
d_f   <- as.integer(d_tab)
d_rf  <- d_f / sum(d_f)
stopifnot(length(drinks) == 20, sum(d_f) == 20,
          identical(d_f, c(6L, 2L, 8L, 4L)),
          names(d_tab)[which.max(d_f)] == "Latte",
          abs(d_rf[3] - 0.40) < 1e-12)
show_set(drinks)
```

**(a)** Complete the frequency and relative frequency table.

```{r p04-table}
show_tbl(data.frame(
  Drink = names(d_tab),
  `Frequency $F(x)$` = cellv(d_f),
  `Relative frequency $RF(x)$` = cellv(sprintf("%.2f", d_rf)),
  check.names = FALSE))
```

**(b)** What is the modal category?

`r ans("Latte, with a frequency of 8")`

**(c)** Draw a bar graph of the distribution on the axes below.

```{r p04-bar, fig.height=3.6, fig.width=5.5}
bar_df <- data.frame(Drink = names(d_tab), Frequency = d_f)
p <- ggplot(bar_df, aes(x = Drink, y = Frequency)) +
  scale_y_continuous(limits = c(0, 9), breaks = 0:9) +
  theme_bw() + xlab("Drink") + ylab("Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p + geom_col(fill = "lightblue", colour = "black", width = 0.65)
} else {
  p + geom_blank()
}
```

**(d)** Would it make sense to add a *cumulative relative frequency* column to the table in
part (a)? Explain.

```{r p04-d, results='asis'}
why(paste("No. Cumulative relative frequency is the proportion of observations",
          "<i>less than or equal to</i> a given value, and that phrase only means",
          "something if the values can be put in order. Drink type is a <b>nominal</b>",
          "variable -- there is no sense in which a mocha is 'less than' a latte -- so",
          "there is no meaningful order to accumulate along.",
          "<br><br>Had the variable been ordinal, say drink size (small, medium, large),",
          "a cumulative relative frequency column would be perfectly sensible."))
workspace(3)
```

---

## Problem 5 — Frequency tables and the mean from a table

Twenty-five households were asked how many vehicles they own. The responses are summarized
below.

```{r p05-data}
v5 <- 0:4
f5 <- c(3, 9, 8, 4, 1)
rf5 <- f5 / sum(f5)
crf5 <- cumsum(rf5)
mean5 <- sum(v5 * f5) / sum(f5)
stopifnot(sum(f5) == 25, mean5 == 1.64,
          abs(crf5[3] - 0.80) < 1e-12,
          abs(sum(v5 * rf5) - mean5) < 1e-12)
show_tbl(data.frame(
  `Vehicles $x$` = v5,
  `Frequency $F(x)$` = f5,
  `Relative frequency $RF(x)$` = cellv(sprintf("%.2f", rf5)),
  `Cumulative relative frequency $CRF(x)$` = cellv(sprintf("%.2f", crf5)),
  check.names = FALSE))
```

**(a)** Complete the relative frequency and cumulative relative frequency columns.

**(b)** Compute the mean number of vehicles using the **frequency** formula,
$\bar{x} = \frac{1}{n}\sum_{x} x\,F(x)$.

```{r p05-b, results='asis'}
why(paste("$\\bar{x} = \\frac{0(3) + 1(9) + 2(8) + 3(4) + 4(1)}{25}",
          "= \\frac{0 + 9 + 16 + 12 + 4}{25} = \\frac{41}{25} = 1.64$ vehicles."))
workspace(3)
```

**(c)** Now compute it again using the **relative frequency** formula,
$\bar{x} = \sum_{x} x\,RF(x)$, and confirm you get the same answer.

```{r p05-c, results='asis'}
why(paste("$\\bar{x} = 0(0.12) + 1(0.36) + 2(0.32) + 3(0.16) + 4(0.04)",
          "= 0 + 0.36 + 0.64 + 0.48 + 0.16 = 1.64$ vehicles.",
          "<br><br>The two formulas are the same calculation. Dividing by $n$ at the end",
          "(the frequency version) and dividing by $n$ at the start to build the relative",
          "frequencies (this version) get you to the same place. This is also why you",
          "must not divide by $n$ <i>again</i> in the relative frequency version."))
workspace(3)
```

**(d)** What proportion of these households own 2 or fewer vehicles? Which column did you
read that from?

`r ans("$0.80$, read from the cumulative relative frequency column at $x = 2$")`

**(e)** What is the median number of vehicles owned? (You can read this off the cumulative
relative frequency column.)

```{r p05-e, results='asis'}
why(paste("The median is the $50^{th}$ percentile. Reading down the $CRF$ column, it",
          "passes $0.50$ at $x = 2$ (the $CRF$ is $0.48$ at $x = 1$ and $0.80$ at",
          "$x = 2$), so the <b>median is 2 vehicles</b>.",
          "<br><br>Check it against the raw data if you like: with $n = 25$ the median is",
          "the $13^{th}$ value in order, and values 13 through 20 are all 2s."))
workspace(3)
```

---

## Problem 6 — Binning a continuous variable

The following are the times, in seconds, that 16 users took to complete a task on a website.

```{r p06-data, results='asis'}
t6 <- c(12.4, 14.1, 15.8, 16.2, 17.5, 18.0, 18.3, 19.1,
        20.2, 20.6, 21.7, 22.4, 23.9, 25.1, 26.8, 28.4)
k6 <- round(sqrt(length(t6)))
w6 <- (max(t6) - min(t6)) / k6
brk6 <- c(12, 16, 20, 24, 28.5)
f6 <- as.integer(table(cut(t6, breaks = brk6, right = TRUE)))
stopifnot(length(t6) == 16, k6 == 4, sum(f6) == 16,
          identical(f6, c(3L, 5L, 5L, 3L)), abs(w6 - 4.0) < 0.01)
show_set(t6)
```

**(a)** Using the **square root rule**, how many bins $k$ should you use for a histogram of
these 16 observations?

`r ans("$k = \\text{round}(\\sqrt{16}) = 4$ bins")`

**(b)** Compute the bin width $w$ implied by that choice, using
$w = \frac{\max x - \min x}{k}$.

`r ans("$w = (28.4 - 12.4)/4 = 16/4 = 4.0$ seconds")`

**(c)** Complete the frequency table using the intervals given.

```{r p06-table}
show_tbl(data.frame(
  `Time (seconds)` = c("$12 < x \\leq 16$", "$16 < x \\leq 20$",
                       "$20 < x \\leq 24$", "$24 < x \\leq 28.5$"),
  `Frequency $F(x)$` = cellv(f6),
  `Relative frequency $RF(x)$` = cellv(sprintf("%.4f", f6 / sum(f6))),
  check.names = FALSE))
```

**(d)** Draw the histogram on the axes below.

```{r p06-hist, fig.height=3.8, fig.width=5.5}
hist_df <- data.frame(mid = c(14, 18, 22, 26.25), f = f6,
                      wid = diff(brk6))
p <- ggplot(hist_df, aes(x = mid, y = f, width = wid)) +
  scale_x_continuous(limits = c(12, 28.5), breaks = c(12, 16, 20, 24, 28.5)) +
  scale_y_continuous(limits = c(0, 6), breaks = 0:6) +
  theme_bw() + xlab("Time to complete task (seconds)") + ylab("Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p + geom_col(fill = "lightblue", colour = "black")
} else {
  p + geom_blank()
}
```

**(e)** Describe the shape of this distribution.

```{r p06-e, results='asis'}
why(paste("Roughly <b>symmetric and unimodal</b>, with the bulk of the observations in",
          "the two middle bins and the counts tapering off evenly at both ends. There is",
          "no long tail in either direction and no obvious gap, so nothing here looks",
          "like an outlier."))
workspace(3)
```

**(f)** Give one thing you can determine from the table in part (c), and one thing you
**cannot**, that you could have determined from the raw data.

```{r p06-f, results='asis'}
why(paste("<b>Can:</b> the proportion of users finishing in 20 seconds or less",
          "($8/16 = 0.5$), or the number in any one interval.",
          "<b>Cannot:</b> the fastest or slowest individual time, the exact mean, or the",
          "median -- once the values are grouped, the individual measurements are gone.",
          "<br><br>That loss of precision is the cost of binning. It buys a readable",
          "summary of the shape, which the raw list of 16 numbers does not give you."))
workspace(3)
```

---

## Problem 7 — Dot plots and stem-and-leaf plots

A biologist recorded the number of eggs in each of 15 birds' nests:

```{r p07-data, results='asis'}
e7 <- c(21, 24, 24, 27, 28, 31, 33, 33, 33, 35, 38, 41, 42, 44, 47)
mode7 <- as.integer(names(sort(table(e7), decreasing = TRUE))[1])
stopifnot(length(e7) == 15, mode7 == 33, identical(e7, sort(e7)))
show_set(e7)
```

**(a)** Construct a **stem-and-leaf plot**, using the tens digit as the stem.

```{r p07-stem, results='asis'}
if (SOLUTIONS) {
  stems <- 2:4
  rows <- vapply(stems, function(s) {
    leaves <- e7[e7 %/% 10 == s] %% 10
    sprintf("  %d | %s", s, paste(leaves, collapse = " "))
  }, character(1))
  cat('<pre style="color: red;">', paste(rows, collapse = "\n"), "</pre>", sep = "\n")
} else {
  cat('<pre>  2 |\n  3 |\n  4 |\n</pre>')
}
```

**(b)** Construct a **dot plot** on the number line below.

```{r p07-dot, fig.height=2.6, fig.width=6.5}
dot_df <- data.frame(x = e7)
dot_df$y <- ave(dot_df$x, dot_df$x, FUN = seq_along)
p <- ggplot(dot_df, aes(x = x, y = y)) +
  scale_x_continuous(limits = c(20, 48), breaks = seq(20, 48, 2)) +
  scale_y_continuous(limits = c(0.5, 4)) +
  theme_classic() + xlab("Eggs per nest") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), axis.line.y = element_blank())
if (SOLUTIONS) p + geom_point(size = 2.6) else p + geom_blank()
```

**(c)** What is the mode of this variable?

`r ans("33 eggs, which occurs three times")`

**(d)** Is the mode a measure of the center of this distribution? Explain.

```{r p07-d, results='asis'}
why(paste("No -- the mode is a measure of <b>location</b>. It tells you where the peak",
          "is, i.e. the most frequent value, not where the center is. Here the",
          "distribution happens to be roughly symmetric so the mode of 33 does land near",
          "the middle, but that is a coincidence of this data rather than a property of",
          "the mode.",
          "<br><br>For a skewed distribution the mode can sit well away from the center,",
          "and a bimodal distribution has two modes with the center usually falling in",
          "the gap between them, where the data are sparsest."))
workspace(3)
```

---

## Problem 8 — Center, and what an outlier does to it

Consider the following eight observations of a variable $X$, the number of minutes eight
customers spent in a shop:

```{r p08-data, results='asis'}
x8 <- c(4, 6, 6, 9, 11, 12, 14, 16)
mean8 <- mean(x8); med8 <- median(x8)
x8b <- c(x8, 78)
mean8b <- mean(x8b); med8b <- median(x8b)
stopifnot(mean8 == 9.75, med8 == 10, mean8b == 17.333 | abs(mean8b - 156/9) < 1e-9,
          med8b == 11, sum(x8) == 78)
show_set(x8)
```

**(a)** Compute the mean, the median and the mode.

```{r p08-a}
show_tbl(data.frame(
  Statistic = c("Mean $\\bar{x}$", "Median", "Mode"),
  `Your answer` = cellv(c("$78/8 = 9.75$ minutes", "$(9+11)/2 = 10$ minutes", "6 minutes")),
  check.names = FALSE))
```

**(b)** A ninth customer stays for **78 minutes**. Recompute the mean and the median with
that observation included.

```{r p08-b, results='asis'}
why(paste("$\\bar{x} = (78 + 78)/9 = 156/9 \\approx 17.33$ minutes.",
          "<br>The median is now the $5^{th}$ of 9 ordered values, which is",
          "<b>11 minutes</b>."))
workspace(3)
```

**(c)** Which statistic moved more, and why?

```{r p08-c, results='asis'}
why(paste("The <b>mean</b> moved far more -- from 9.75 up to about 17.33, an increase of",
          "roughly 7.6 minutes -- while the median moved from 10 to 11, just one minute.",
          "<br><br>The mean uses the <i>value</i> of every observation, so a single very",
          "large value pulls it upward in proportion to how extreme it is. The median",
          "depends only on <i>position</i> in the ordered data; adding one more value at",
          "the top shifts the middle position over by one place and nothing more.",
          "<br><br>Notice too that after the outlier the mean of 17.33 is larger than",
          "<i>every observation except the outlier itself</i>. It is no longer describing",
          "a typical customer at all."))
workspace(4)
```

**(d)** Which of the two is **resistant**? Give a one-sentence definition of that word.

`r ans("The median is resistant.")`

```{r p08-d, results='asis'}
why(paste("A summary measure is <b>resistant</b> when a single extreme observation barely",
          "moves it. Whether a measure is resistant comes down to one thing: whether it",
          "uses the values of every observation, or only positions in the ordered data."))
workspace(2)
```

---

## Problem 9 — Percentiles and quartiles

The following are the ages of the 11 members of a community board, sorted:

```{r p09-data, results='asis'}
a9 <- c(29, 34, 38, 41, 45, 49, 52, 56, 61, 67, 73)
q9 <- q_halves(a9)
## Part (c) adds a 12th member, so the same data is worked once with n odd (drop
## the median) and once with n even (no median to drop) -- the two halves of the
## rule, on one dataset.
a9b <- sort(c(a9, 58))
q9b <- q_halves(a9b)
stopifnot(length(a9) == 11, identical(a9, sort(a9)),
          q9[["Q1"]] == 38, q9[["Q2"]] == 49, q9[["Q3"]] == 61,
          length(a9b) == 12,
          q9b[["Q1"]] == 39.5, q9b[["Q2"]] == 50.5, q9b[["Q3"]] == 59.5)
show_set(a9)
```

**(a)** Find the median, $Q2$.

`r ans("$Q2 = 49$, the $6^{th}$ of 11 ordered values")`

**(b)** Find $Q1$ and $Q3$ using the **method of halves** we use in this class. Remember
that when $n$ is odd the median is **left out** of both halves.

```{r p09-b, results='asis'}
why(paste("Dropping the median, the lower half is $\\{29, 34, 38, 41, 45\\}$ and the",
          "upper half is $\\{52, 56, 61, 67, 73\\}$. Each has five values, so each",
          "median is its middle value:",
          "<br><br>$Q1 = 38$ years and $Q3 = 61$ years."))
workspace(4)
```

**(c)** A twelfth member, aged $58$, joins the board. Find $Q1$, $Q2$ and $Q3$ for the
new set of 12 ages.

```{r p09-c, results='asis'}
why(paste("The ordered ages are now",
          "$\\{29, 34, 38, 41, 45, 49, 52, 56, 58, 61, 67, 73\\}$.",
          "<br><br>With $n = 12$ there is no single middle value, so the median is the",
          "average of the $6^{th}$ and $7^{th}$ ages: $Q2 = (49 + 52)/2 = 50.5$ years.",
          "Because $n$ is even there is nothing to leave out -- the data splits cleanly",
          "into two halves of six.",
          "<br><br>The lower half is $\\{29, 34, 38, 41, 45, 49\\}$, so",
          "$Q1 = (38 + 41)/2 = 39.5$ years.",
          "<br>The upper half is $\\{52, 56, 58, 61, 67, 73\\}$, so",
          "$Q3 = (58 + 61)/2 = 59.5$ years.",
          "<br><br>Compare this with part (b): the method is the same both times. The",
          "only thing that changes is that an odd $n$ leaves you a middle value to drop,",
          "and an even $n$ does not."))
workspace(5)
```

**(d)** What is the $IQR$, and what does it measure?

`r ans("$IQR = 61 - 38 = 23$ years")`

```{r p09-d, results='asis'}
why(paste("The IQR is the distance between $Q1$ and $Q3$, so it describes how spread out",
          "the <b>middle 50%</b> of the data are. Because it ignores the extreme ends of",
          "the distribution entirely, it is highly resistant to outliers."))
workspace(2)
```

**(e)** The board member aged 67 is at roughly what percentile? Answer in a sentence that
says what that means.

```{r p09-e, results='asis'}
why(paste("Ten of the 11 members are aged 67 or younger, so 67 sits at about the",
          "$10/11 \\approx 91^{st}$ percentile: roughly 91% of the board is at or below",
          "that age, and only one member is older."))
workspace(3)
```

---

## Problem 10 — Deviations, variance and standard deviation

A quality inspector measured the weight, in grams, of six packets from a filling machine:

```{r p10-data, results='asis'}
w10 <- c(48, 51, 52, 55, 57, 59)
m10 <- mean(w10)
dev10 <- w10 - m10
ss10 <- sum(dev10^2)
var10 <- ss10 / (length(w10) - 1)
sd10 <- sqrt(var10)
stopifnot(abs(m10 - 322/6) < 1e-9, sum(w10) == 322,
          abs(sum(dev10)) < 1e-12, abs(ss10 - 83.3333) < 1e-3,
          abs(var10 - 16.6667) < 1e-3, abs(sd10 - 4.0825) < 1e-3)
show_set(w10)
```

**(a)** Compute the sample mean $\bar{x}$.

`r ans("$\\bar{x} = 322/6 \\approx 53.67$ grams")`

**(b)** Complete the deviation table below, then add up the deviation column.

```{r p10-table}
show_tbl(data.frame(
  `$x_i$` = w10,
  `$x_i - \\bar{x}$` = cellv(sprintf("%.2f", dev10)),
  `$(x_i - \\bar{x})^2$` = cellv(sprintf("%.2f", dev10^2)),
  check.names = FALSE))
```

```{r p10-b, results='asis'}
why(paste("The deviation column sums to <b>0</b> (up to rounding), and the squared",
          "deviation column sums to about <b>83.33</b>.",
          "<br><br>The deviations summing to zero is not a property of these six numbers",
          "-- it happens for every dataset, because the mean is the balancing point. That",
          "is exactly why the variance squares the deviations before adding them: without",
          "squaring, the positives and negatives always cancel and you learn nothing."))
workspace(3)
```

**(c)** Compute the sample variance $s^2$ and the sample standard deviation $s$.

```{r p10-c, results='asis'}
why(paste("$s^2 = \\frac{83.33}{6 - 1} = \\frac{83.33}{5} \\approx 16.67$",
          "<br><br>$s = \\sqrt{16.67} \\approx 4.08$ grams"))
workspace(3)
```

**(d)** State the units of $\bar{x}$, of $s^2$ and of $s$.

```{r p10-d}
show_tbl(data.frame(
  Quantity = c("$\\bar{x}$", "$s^2$", "$s$"),
  Units = cellv(c("grams", "grams squared", "grams")),
  check.names = FALSE))
```

```{r p10-d2, results='asis'}
why(paste("Squaring the deviations squares their units too, which is why the variance is",
          "in grams squared -- a quantity nobody can picture. Taking the square root",
          "undoes it and puts $s$ back on the scale of the original data, where it reads",
          "as roughly the typical distance of a packet from the mean weight."))
workspace(2)
```

**(e)** Compute the range. Which is the better description of the spread of this data, the
range or the standard deviation, and why?

```{r p10-e, results='asis'}
why(paste("Range $= 59 - 48 = 11$ grams.",
          "<br><br>The <b>standard deviation</b> is the better description here. The range",
          "is built from the two most extreme observations and <i>nothing else</i>, so it",
          "throws away the other four measurements entirely and moves by the full amount",
          "if either endpoint changes. The standard deviation uses every observation.",
          "<br><br>The range is worth reporting as a quick descriptive fact, but it is",
          "almost never used as <i>the</i> measure of spread."))
workspace(3)
```

---

## Problem 11 — Five-number summary, boxplots and outliers

A clinic recorded the number of days between a referral and an appointment for 12 patients:

```{r p11-data, results='asis'}
d11 <- c(3, 5, 6, 8, 9, 11, 12, 14, 15, 17, 19, 58)
q11 <- q_halves(d11)
iqr11 <- q11[["Q3"]] - q11[["Q1"]]
lo11 <- q11[["Q1"]] - 1.5 * iqr11
hi11 <- q11[["Q3"]] + 1.5 * iqr11
out11 <- d11[d11 < lo11 | d11 > hi11]
whisk_hi <- max(d11[d11 <= hi11])

## n is EVEN on purpose. base boxplot() draws Tukey's hinges, which agree with
## the class's median-of-halves rule only when n is even -- with an odd n the
## drawn box would contradict the five-number summary in part (a). Asserted
## rather than trusted.
stopifnot(length(d11) == 12, identical(d11, sort(d11)),
          all.equal(as.numeric(fivenum(d11)[c(2, 4)]),
                    c(q11[["Q1"]], q11[["Q3"]])),
          q11[["Q1"]] == 7, q11[["Q2"]] == 11.5, q11[["Q3"]] == 16,
          iqr11 == 9, lo11 == -6.5, hi11 == 29.5,
          identical(out11, 58), whisk_hi == 19,
          abs(mean(d11) - 14.75) < 1e-9)
show_set(d11)
```

**(a)** Give the five-number summary. Use the method of halves for the quartiles.

```{r p11-a}
show_tbl(data.frame(
  Minimum = cellv("3"), Q1 = cellv("7"), Median = cellv("11.5"),
  Q3 = cellv("16"), Maximum = cellv("58"), check.names = FALSE))
```

```{r p11-a2, results='asis'}
why(paste("$n = 12$ is even, so there is no middle value to leave out: the median is the",
          "average of the $6^{th}$ and $7^{th}$ values, $Q2 = (11+12)/2 = 11.5$ days, and",
          "the data splits cleanly into two halves of six.",
          "<br><br>The lower half is $\\{3,5,6,8,9,11\\}$ with median",
          "$Q1 = (6+8)/2 = 7$; the upper half is $\\{12,14,15,17,19,58\\}$ with median",
          "$Q3 = (15+17)/2 = 16$."))
workspace(3)
```

**(b)** Compute the $IQR$ and both outlier boundaries.

```{r p11-b, results='asis'}
why(paste("$IQR = 16 - 7 = 9$ days, so $1.5 \\times IQR = 13.5$.",
          "<br><br>Lower boundary $= 7 - 13.5 = -6.5$ days.",
          "<br>Upper boundary $= 16 + 13.5 = 29.5$ days.",
          "<br><br>Nothing can fall below $-6.5$ days, so only the upper boundary can",
          "flag anything here."))
workspace(3)
```

**(c)** Which observations, if any, does the $1.5 \times IQR$ rule flag as outliers?

`r ans("Just one: the 58-day wait, which is above the upper boundary of 29.5")`

**(d)** Draw the boxplot on the axis below. Be careful about where the right whisker ends.

```{r p11-box, fig.height=2.8, fig.width=6.5}
if (SOLUTIONS) {
  par(mar = c(4, 1, 1, 1))
  boxplot(d11, horizontal = TRUE, axes = FALSE, staplewex = 1, col = "lightgrey",
          ylim = c(0, 60))
  axis(1, at = seq(0, 60, 5))
  mtext("Days from referral to appointment", side = 1, line = 2.5)
} else {
  par(mar = c(4, 1, 1, 1))
  plot(NA, xlim = c(0, 60), ylim = c(0.5, 1.5), axes = FALSE, xlab = "", ylab = "")
  axis(1, at = seq(0, 60, 5))
  mtext("Days from referral to appointment", side = 1, line = 2.5)
}
```

```{r p11-d, results='asis'}
why(paste("The right whisker stops at <b>19</b>, the largest observation that is still at",
          "or below the upper boundary of 29.5 -- <i>not</i> at the maximum of 58, and",
          "not at the boundary itself. A whisker always ends on an actual data value. The",
          "58 is drawn as its own point beyond the whisker.",
          "<br><br>The left whisker runs all the way out to the minimum of 3, because",
          "nothing is flagged at the low end."))
workspace(3)
```

**(e)** Describe the shape of this distribution from the boxplot, and say what that implies
about the mean relative to the median.

```{r p11-e, results='asis'}
why(paste("The distribution is <b>skewed right</b>: the median sits left of centre in the",
          "box, the right whisker is longer than the left, and there is a high outlier.",
          "<br><br>A right skew pulls the mean toward the long tail, so the",
          "<b>mean will be greater than the median</b>. The mean here is 14.75 days",
          "against a median of 11.5."))
workspace(3)
```

**(f)** The clinic manager wants to delete the 58-day wait because it is "clearly a
mistake." What would you tell her?

```{r p11-f, results='asis'}
why(paste("That being flagged by the $1.5 \\times IQR$ rule is an <b>arithmetic result,",
          "not a verdict</b>. The rule reports only that a value sits unusually far from",
          "the middle half of the data; it cannot distinguish a recording error from a",
          "genuinely unusual case, and it marks the observation as worth a second look",
          "rather than as wrong.",
          "<br><br>A 58-day wait is entirely plausible -- it may be the single most",
          "important number in the dataset for a manager trying to improve scheduling.",
          "Deleting it because a rule flagged it would throw away real information."))
workspace(4)
```

---

## Problem 12 — Cumulative distributions

A librarian recorded the number of books checked out by each of 20 patrons in one hour.

```{r p12-data}
x12 <- 0:5
f12 <- c(2, 5, 6, 4, 2, 1)
rf12 <- f12 / sum(f12)
crf12 <- cumsum(rf12)
tab12 <- data.frame(x = x12, crf = crf12)
raw12 <- rep(x12, f12)
q12 <- q_halves(raw12)
stopifnot(sum(f12) == 20, length(raw12) == 20,
          abs(crf12[3] - 0.65) < 1e-12,
          q12[["Q1"]] == 1, q12[["Q2"]] == 2, q12[["Q3"]] == 3,
          sum(x12 * f12) == 42, mean(raw12) == 2.1)
```

**(a)** Complete the table.

```{r p12-table}
show_tbl(data.frame(
  `Books $x$` = x12,
  `Frequency $F(x)$` = f12,
  `Relative frequency $RF(x)$` = cellv(sprintf("%.2f", rf12)),
  `Cumulative relative frequency $CRF(x)$` = cellv(sprintf("%.2f", crf12)),
  check.names = FALSE))
```

**(b)** Plot the cumulative distribution on the axes below. Remember it is a **step
function**: flat between observed values, jumping at each one.

```{r p12-cumdist, fig.height=4.6, fig.width=7}
lead12 <- data.frame(x = -1, crf = 0)
p <- ggplot() +
  scale_x_continuous(limits = c(-1, 6), breaks = 0:6) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.05)) +
  theme_bw() + xlab("Books checked out") + ylab("Cumulative Relative Frequency") +
  theme(panel.grid.major.y = element_line(colour = "grey70"))
if (SOLUTIONS) {
  p +
    geom_step(data = rbind(lead12, tab12), aes(x = x, y = crf),
              direction = "hv", linewidth = 1) +
    geom_point(data = tab12, aes(x = x, y = crf), size = 3)
} else {
  p + geom_blank()
}
```

**(c)** Use the plot to find $Q1$, the median, and $Q3$.

```{r p12-c, results='asis'}
why(paste("Read across from $0.25$, $0.50$ and $0.75$ on the $y$-axis to the curve, then",
          "drop to the $x$-axis. Each of the three readings crosses a jump:",
          "<br><br>$Q1 = 1$ book, $Q2 = 2$ books, $Q3 = 3$ books.",
          "<br><br>Check them against halving the ordered data -- they agree. The",
          "cumulative distribution is not a new definition of a quartile, just a faster",
          "way to read one off, and unlike halving it gives you <i>any</i> percentile."))
workspace(4)
```

**(d)** What proportion of patrons checked out 2 or fewer books?

`r ans("$0.65$, read straight off the $CRF$ column at $x = 2$")`

**(e)** Describe the shape of this distribution, and say whether the mean will be greater
than, less than, or equal to the median.

```{r p12-e, results='asis'}
why(paste("The distribution is <b>skewed right</b>: the frequencies rise to a peak at",
          "2 books and then trail off toward the larger values.",
          "<br><br>A right skew pulls the mean toward the long tail while leaving the",
          "median where it is, so the <b>mean is greater than the median</b>. Here the",
          "mean is $42/20 = 2.1$ books against a median of 2."))
workspace(3)
```

---

```{r review-outro, results='asis'}
if (!SOLUTIONS) {
  cat(paste(
    "## Check your work\n\n",
    "Worked solutions are posted under",
    "**Exam Study Guides &rarr; Exam 1 Review &rarr; Solutions**. Work each problem",
    "through before looking at them.\n\n",
    "Every quartile on this exam uses the method of halves. Problem 9 works it both",
    "ways: with an odd $n$, where the median is dropped, and with an even $n$, where",
    "it is not.\n"))
} else {
  cat(paste(
    "## A note on the exam\n\n",
    "These solutions do more arithmetic than the exam requires. Exam 1 asks you to",
    "interpret values, identify equations and read plots. The formulas are shown during",
    "the exam, but **unlabeled** -- identifying which is which is on you.\n"))
}
```
