---
title: "Significance Testing R Functions"
author: "STAT 251 Section 01"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4)

# The code shown on this page is NOT typed out here. It is pulled straight from
# misc/stat251_tools.R by read_chunk(), which splits that file on its
# `## ---- name ----` marker lines and hands each block back under that name.
# An empty chunk below like ```{r one-sample-t-test, eval=FALSE}``` prints the
# matching block. That way this page can never disagree with the file the
# lectures, homework solutions and exams actually source().
knitr::read_chunk('stat251_tools.R')

# ...and source it for real, so the worked examples below actually run.
source('stat251_tools.R')

library(knitr)
library(kableExtra)

# Download link with the file size filled in automatically, same helper style
# as homeworks/homework_downloads.Rmd. The `download` attribute makes the
# browser save the file instead of displaying it as text.
dl <- function(path, label = basename(path)) {
  kb <- if (file.exists(path)) sprintf(" (%.0f KB)", max(file.size(path)/1024, 1)) else ""
  sprintf('<a href="%s" download>%s</a>%s', path, label, kb)
}
```

This page is a directory of the R functions written for STAT 251. They cover the
confidence intervals and significance tests we build by hand in class, and they
print their results in the same order we write them on the board: hypotheses,
estimate, standard error, critical value, interval, test statistic, p-value,
decision.

You do **not** need any of this to do well in the course — everything here can be
done with a calculator and a table. It is here if you want to check your
arithmetic, or if you are curious how these calculations look in code.

## Getting the functions

Every function below lives in one file:

```{r echo=FALSE, results='asis'}
cat("- ", dl("stat251_tools.R"), " — download this into your working folder\n", sep = "")
```

Save it next to your R script, then load all of it in one line:

```{r eval=FALSE}
source('stat251_tools.R')
```

That is it — every function on this page is now available. If R replies
`cannot open file 'stat251_tools.R'`, R is looking in a different folder than the
one you saved it to. In RStudio, use **Session → Set Working Directory → To
Source File Location** and run the line again.

A few conventions shared by all of them:

* `alpha` is the significance level, so `alpha = 0.05` gives a 95% interval.
* `test` picks the alternative hypothesis: `'lower.tail'`, `'upper.tail'` or
  `'two.tail'`.
* `verbose = TRUE` prints the full worked result. The test functions default to
  `TRUE`; the interval functions default to `FALSE` and just return the numbers.
* The functions take **summary statistics** (a mean, a standard deviation, a
  count), not raw data columns — the same numbers a problem gives you. The
  non-parametric tests are the exception, since ranks need the actual values.

<br>

# Picturing the test

Before the arithmetic, it helps to see what is being asked. `gen.density.plot()`
draws the distribution of the test statistic when $H_0$ is true, shades the
rejection region, and marks both the critical value and the statistic you
observed. If the observed point lands in the shaded area, you reject.

```{r density-plot, eval=FALSE}
```

`d_limit()` is its helper. It returns the height of the curve, but `NA` for every
x outside the rejection region — and since `ggplot` silently skips `NA`s, the
shading lands only on the tail(s).

```{r rejection-region-helper, eval=FALSE}
```

**Example.** A one-tailed z test at $\alpha = 0.05$ with an observed statistic of
2.1. The critical value is 1.645, the observed value is past it, so this is a
rejection:

```{r}
gen.density.plot(dist = 'z', alpha = 0.05, obs = 2.1, test = 'upper.tail')
```

The same picture for a t test with 15 observations — note the heavier tails,
which push the critical value further out than the z:

```{r}
gen.density.plot(dist = 't', n = 15, alpha = 0.05, obs = 2.1, test = 'two.tailed')
```

<br>

# One sample: a proportion

Use these when the question is about a **percentage or a fraction** of one group
— what share of students own a car, whether a coin is fair.

## Standard error

```{r one-sample-prop-se, eval=FALSE}
```

## Confidence interval

```{r one-sample-prop-ci, eval=FALSE}
```

## Test

```{r one-sample-prop-test, eval=FALSE}
```

**Example.** In a sample of 100 students, 62 said they own a car. Is that
evidence that more than half of all students own one?

```{r}
one.sample.prop.test(p0 = 0.5, phat = 0.62, n = 100, test = 'upper.tail')
```

The p-value is 0.0082, well under 0.05, so we reject $H_0$.

Note that the standard error printed here (0.05) is computed from $p_0 = 0.5$,
not from $\hat p = 0.62$. While we are testing, we assume $H_0$ is true. The
confidence interval makes the opposite choice — it has no null value to lean on,
so it uses $\hat p$:

```{r}
one.sample.prop.CI(phat = 0.62, n = 100, alpha = 0.05, verbose = TRUE)
```

<br>

# One sample: a mean

Use these when the question is about an **average** of one group — mean study
time, mean height — and the population standard deviation is unknown, which in
practice is always.

## Standard error

```{r one-sample-t-se, eval=FALSE}
```

## Confidence interval

```{r one-sample-t-ci, eval=FALSE}
```

## Test

```{r one-sample-t-test, eval=FALSE}
```

**Example.** 25 students studied a mean of 13.2 hours a week with a standard
deviation of 5.4 hours. Is the true mean different from 12?

```{r}
one.sample.t.test(m0 = 12, xbar = 13.2, s = 5.4, n = 25, test = 'two.tail')
```

The p-value is 0.2775, so we fail to reject $H_0$. Consistent with that, the 95%
interval contains 12:

```{r}
one.sample.t.CI(xbar = 13.2, s = 5.4, n = 25, alpha = 0.05, verbose = TRUE)
```

<br>

# Two samples: proportions

Use these to **compare a percentage between two groups**.

## Standard error

```{r two-sample-prop-se, eval=FALSE}
```

## Confidence interval

```{r two-sample-prop-ci, eval=FALSE}
```

## Test

```{r two-sample-prop-test, eval=FALSE}
```

**Example.** 45 of 80 students in one group and 30 of 90 in another answered
yes. Note this function wants the **counts**, not the proportions:

```{r}
two.sample.prop.test(p0 = 0, x1 = 45, n1 = 80, x2 = 30, n2 = 90, test = 'two.tail')
```

`pooled = TRUE` is the default and is usually what you want: $H_0$ says the two
proportions are equal, so under $H_0$ there is only one proportion to estimate
and both samples can be pooled into it. The interval is always built from the
*unpooled* standard error, because an interval does not assume $H_0$ is true.

<br>

# Two samples: means

Use these to **compare an average between two groups**.

## Standard error

```{r two-sample-t-se, eval=FALSE}
```

## Confidence interval

```{r two-sample-t-ci, eval=FALSE}
```

## Test

```{r two-sample-t-test, eval=FALSE}
```

**Example.** Group 1: mean 13.2, sd 5.4, n = 25. Group 2: mean 10.8, sd 4.9,
n = 30.

```{r}
two.sample.t.test(m0 = 0, x1 = 13.2, x2 = 10.8, s1 = 5.4, s2 = 4.9,
                  n1 = 25, n2 = 30, test = 'two.tail', pooling = 'unpooled')
```

The `pooling` argument chooses how to handle the two standard deviations:

* `'pooled'` — assume the two populations have equal variances. Combines them
  into one estimate and gets the full $n_1 + n_2 - 2$ degrees of freedom.
* `'unpooled'` — do not assume that. Uses the Welch–Satterthwaite formula, which
  usually gives a fractional degrees of freedom (49.1 above). That is normal.
* `'approx.unpooled'` — the same standard error with the easy hand-calculation
  shortcut $df = \min(n_1, n_2) - 1$, which is what we use on paper.

Switching to the pooled version on the same data barely moves the answer, which
is typical when the two sample standard deviations are close:

```{r}
two.sample.t.test(m0 = 0, x1 = 13.2, x2 = 10.8, s1 = 5.4, s2 = 4.9,
                  n1 = 25, n2 = 30, test = 'two.tail', pooling = 'pooled')
```

<br>

# Non-parametric tests

These make no assumption that the data are normal. They work on **signs and
ranks** instead of the actual measurements, which costs some power but survives
outliers and small samples. Unlike the functions above, these need the raw data.

## The sign test

Throws away how big each difference is and keeps only whether it was positive or
negative, then tests those signs as a proportion. Pairs that tie exactly are
dropped, so the number of signs can be smaller than the number of pairs.

```{r sign-test, eval=FALSE}
```

**Example.** Ten students' resting heart rate before and after a program:

```{r}
before <- c(72, 68, 75, 80, 66, 71, 79, 74, 70, 77)
after  <- c(70, 69, 71, 74, 65, 68, 73, 74, 66, 72)

sign.test(p0 = 0.5, x1 = before, x2 = after, test = 'upper.tail')
```

Eight of the ten differences were positive; one student tied and was dropped,
leaving 9 signs. The p-value of 0.0195 comes from the exact binomial, not a
normal approximation.

## Wilcoxon rank-sum test (independent samples)

The non-parametric stand-in for the two-sample t test. Pool both samples, rank
everything from smallest to largest, then add up the ranks belonging to the
first sample.

```{r rank-sum-stat, eval=FALSE}
```

```{r wilcoxon-rank-sum-test, eval=FALSE}
```

**Example.** Two independent groups, where group X reads consistently lower:

```{r}
X <- c(11, 13, 14, 16, 18)
Y <- c(19, 21, 22, 25, 27, 30)

rank.sum.stat(X, Y)
```

X holds ranks 1 through 5, so its rank sum is the smallest it could possibly be:
$1+2+3+4+5 = 15$. Testing that:

```{r}
Wilcoxon.rank.sum.test(m0 = 0, X = X, Y = Y, test = 'lower.tail')
```

## Wilcoxon signed-rank test (paired samples)

The non-parametric stand-in for the paired t test. Unlike the sign test it keeps
some information about *size*: rank the absolute differences, then put each
difference's sign back onto its rank.

```{r sign-rank-stat, eval=FALSE}
```

```{r wilcoxon-sign-rank-test, eval=FALSE}
```

**Example.** The same heart-rate data. `sign.rank.stat()` returns every
intermediate step, so you can check a hand calculation against it:

```{r}
sign.rank.stat(before, after)
```

Reading that output: the tied pair is dropped from `absolute.diff`, the ranks
average across ties (two differences of 1 share ranks 1 and 2, so both get 1.5),
and only the one negative difference keeps a minus sign. Adding the signed ranks
gives $W = 42$.

```{r}
Wilcoxon.sign.rank.test(m0 = 0, X = before, Y = after, test = 'upper.tail')
```

This one uses a normal approximation to $W$, so it wants a reasonable number of
pairs before you trust it.

<br>

# Chi-square tests

Both of these compare **counts you observed** against **counts you would expect**,
using the same distance measure — $(\text{observed} - \text{expected})^2 /
\text{expected}$, summed over every cell.

## Goodness of fit

One categorical variable against a claimed set of probabilities.

```{r chi-squared-gof-test, eval=FALSE}
```

**Example.** 100 students picked one of four options. If every option were
equally likely we would expect 25 each:

```{r}
chi.squared.GOF.test(observed.ct  = c(30, 20, 25, 25),
                     expected.freq = rep(0.25, 4))
```

The printed table shows each category's contribution to the statistic, so you can
see which categories are driving the result. Here the total is only 2.0 against a
critical value of 7.81 — no evidence against equal probabilities.

You can give the expected values as counts instead of probabilities, with
`expected.ct`; supply one or the other, not both.

## Independence and homogeneity

Two categorical variables in a two-way table. The arithmetic is identical for
both tests — what differs is how the data were collected, and so how you word the
conclusion.

```{r chi-squared-test, eval=FALSE}
```

**Example.** Car ownership against class standing. Give the table row and column
names and they are used in the printed output:

```{r}
cars <- matrix(c(35, 25,
                 20, 40),
               nrow = 2, byrow = TRUE,
               dimnames = list(c("Yes", "No"), c("Freshman", "Senior")))

chi.squared.test(cars, var1.name = "Class", var2.name = "Owns a car",
                 test = 'independence')
```

The function prints all three tables — observed, expected, and the distances —
so every step of the hand calculation can be checked. With $df = 1$ it also
prints the equivalent z statistic, since a chi-square test on a 2×2 table and a
two-sample z test for proportions are the same test.
