---
title: "R Code Examples"
author: "STAT 251 Section 01"
date: "2026-09-11"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(latex2exp)
library(DT)

path = '../Data/'

# Paginated data preview. Lifted from the make_preview_dt() helper in data.Rmd
# so the two previews behave the same way.
#
# Styling for the DOWNLOAD button, the dropdown and the table header lives in
# assets/brand/site.css - it must NOT be set here as inline style, because
# inline style beats the stylesheet and so cannot follow the dark/light theme.
make_preview_dt <- function(data, label) {
  datatable(
    as.data.frame(data), rownames = TRUE,
    extensions = 'Buttons',
    options = list(
      searching  = FALSE,
      paging     = TRUE,
      bInfo      = TRUE,
      pageLength = 10,
      scrollX    = TRUE,
      columnDefs = list(list(className = 'dt-left',   targets = 0),
                        list(className = 'dt-center', targets = 1:ncol(data))),
      dom = 'Brtip',
      buttons = list(
        list(extend = 'collection', buttons = c('excel', 'csv'), text = label)
      )
    )
  )
}

# Download link with the file size filled in automatically, same helper style as
# homeworks/homework_downloads.Rmd. The `download` attribute makes the browser
# save the file instead of displaying it as text.
dl <- function(p, label = basename(p)) {
  kb <- if (file.exists(p)) sprintf(" (%.0f KB)", max(file.size(p)/1024, 1)) else ""
  sprintf('<a href="%s" download>%s</a>%s', p, label, kb)
}
```

This page shows how to build the graphs and descriptive statistics we cover in
class using R. Everything is done twice where it makes sense: once with the
plotting tools built into R, and once with the **ggplot2** package.

Neither is more correct than the other. Base R is fewer keystrokes for a quick
look at your data. ggplot2 takes longer to learn but gives you far more control
once you know it, and is what most published figures are made with.

# The data

The examples all use a survey of college students in Georgia. Scroll the table
sideways to see all 17 variables, or use the download button to open it in Excel
or Google Sheets:

```{r, echo=FALSE}
gss = read.csv(paste0(path, 'georgia_student_survey.csv'))
make_preview_dt(gss, 'DOWNLOAD: GA Student Survey')
```

```{r echo=FALSE, results='asis'}
cat("You can also download the raw file directly: ",
    dl(paste0(path, 'georgia_student_survey.csv')), "\n", sep = "")
```

**Variable Description**

 * Height - height measured in inches

 * Gender - Gender recorded as female = 0, male = 1

 * Haircut - Cost (in whole dollars) willing to spend on a haircut

 * Job - If the student is employed (Yes = 1, No = 0)

 * Studytime - Average time spent studying over a week (recorded in hours)

 * Smokecig - If the student smokes recorded as No = 0 and Yes = 1

 * Dated - Number of romantic dates in a typical week

 * HSGPA - High school Grade Point Average (GPA)

 * CGPA - College Grade Point Average (GPA)

 * HomeDist - Distance from home to collage (in miles)

 * BrowseInternet - Time spent each day browsing the web (recorded in minutes)

 * WatchTV - Time spent watching TV each day (recorded in minutes)

 * Exercise - Average time spent exercising over a week (recorded in hours)

 * ReadNewsP - Number of days a week the student reads the newspaper

 * Vegan - Whether the student is a vegan (Yes = 1, No = 0)

 * PoliticalDegree - Degree of conservatism or liberalism (1 = extremely liberal, 2 = liberal, 3 = slightly liberal, 4 = moderate, 5 = slightly conservative, 6 = conservative, 7 = extremely conservative)

 * PoliticalAff - The self reported political affilation of the student (1 = Democrat, 2 = Republican, 3 = Democrat)

<br>

# Getting to know a dataset

The first two things to run on any new dataset are `str()` and `summary()`.

`str()` reports the **type** R has given each variable. There are four types
worth knowing about:

* **character** — a qualitative variable stored as text.
* **factor** — a qualitative variable stored as categories, often recorded with
  numbers like 0 and 1.
* **integer** — a quantitative *discrete* variable, i.e. whole numbers.
* **numeric** — a quantitative *continuous* variable.

```{r}
# structure: one line per variable, showing its type and first few values
str(gss)
```

`summary()` gives a five-number summary of every variable at once. Be careful
reading it: a qualitative variable that happens to be recorded as 0/1 will get a
mean and quartiles like any other number, and those are meaningless until you
tell R it is a category.

```{r}
# five-number summary of every column
summary(gss)
```

<br>

# Describing a qualitative variable

We will use **Gender**, recorded as 0 and 1.

R has stored it as numbers, so we first have to convert it to a *factor*. That is
what tells R to treat 0 and 1 as two categories rather than two values to do
arithmetic with.

```{r}
# convert to a factor so R treats the values as categories, not numbers
gss$Gender = as.factor(gss$Gender)

# now reported as a factor with two levels
str(gss$Gender)

# on a factor, summary() counts observations per level instead of averaging
summary(gss$Gender)
```

## Building a frequency table

Both a pie chart and a bar chart need the categories and their proportions, so
we build a small frequency table first.

```{r}
# frequency = the count in each category
freq = summary(gss$Gender)

# relative frequency = each count divided by the sample size
relative.freq = round(freq/sum(freq), 2)

# repackage as a two-column table for plotting
ft = cbind.data.frame(Gender = c('Female', 'Male'), Proportion = relative.freq)
print(ft)
```

## Pie and bar charts in base R

Use `?pie` and `?barplot` to read the documentation for these and see every
argument they accept.

```{r fig.height=3, fig.width=4}
pie(ft$Proportion,        # first argument: non-negative values (counts or proportions)
    labels = ft$Gender,   # category names
    col = c('red','orange'))
```

```{r fig.height=3, fig.width=4}
barplot(height = ft$Proportion,   # first argument: the bar heights
        legend = ft$Gender,       # legend labels
        col = c('red','orange'),
        # move the legend so it does not sit on top of a bar
        args.legend = list(x = 1, y = 1, bty = "n"))
```

## The same charts in ggplot2

Every ggplot is assembled by **adding** pieces together with `+`. You start with
a blank frame from `ggplot()` and add the geometry you want:

`geom_bar()`, `geom_boxplot()`, `geom_histogram()`, `geom_point()`,
`geom_line()`, `geom_dotplot()`, `geom_step()`, and many more — most of which can
be combined.

Every plot also needs an `aes()` argument, the *aesthetic mapping*. That is what
tells ggplot which variable is x, which is y, and which controls colour or size.

ggplot2 has no pie chart. You make one by drawing a stacked bar and then bending
it around a circle with `coord_polar()`:

```{r fig.height=3, fig.width=4}
pie1 = ggplot(ft, aes(x = "", y = Proportion, fill = Gender)) +
  # geom_col is a bar whose height IS the value, rather than a count
  geom_col(color = "black") +
  # print the proportion on each slice
  geom_text(aes(label = Proportion),
            position = position_stack(vjust = 0.6), size = 5) +
  # bend the stacked bar into a circle
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = 'Reds') +
  # theme_void removes all axes and gridlines
  theme_void() +
  theme(legend.text = element_text(size = 12)) +
  ggtitle('Gender')

plot(pie1)
```

The bar chart uses the same frequency table:

```{r fig.height=3, fig.width=4}
BP1 = ggplot(data = ft, aes(x = Gender, y = Proportion, fill = Gender)) +
  # stat = 'identity' means "use the value I gave you", not "count the rows"
  geom_bar(stat = 'identity', color = 'black', linewidth = 0.8) +
  theme_hc() +
  # tick marks every 0.1 from 0 to 0.5
  scale_y_continuous(breaks = seq(0, 0.5, 0.1)) +
  scale_fill_brewer(palette = 'Reds') +
  theme(legend.position = 'top') +
  # theme() controls text, margins and the plot area generally
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  ggtitle("Gender")

plot(BP1)
```

<br>

# Describing a quantitative variable

Now we switch to **Studytime**, the hours per week each student reported
studying.

## Stem plot

```{r}
# look at the raw values first - they are recorded to the nearest half hour
print(gss$Studytime)
```

```{r}
# stem() picks the number of bins for you
stem(gss$Studytime)
```

By default `stem()` guesses the "prettiest" number of bins. The `scale` argument
overrides that guess. Doubling it here stops the rounding to even values that
you can see above:

```{r}
stem(gss$Studytime, scale = 2)
```

## Dot plot

Base R has no good dot plot, so this one is ggplot2 only.

```{r fig.height=4, fig.width=5}
ggplot(data = gss, aes(x = Studytime)) +
  geom_dotplot(dotsize = 0.5) +
  theme_classic() +
  ylab("") +
  xlab("Study Time (hrs)") +
  ggtitle('Dotplot of Student Study Time (Hrs)',
          subtitle = 'Georgia Student Survey') +
  # one tick per hour across the full range of the data
  scale_x_continuous(limits = c(min(gss$Studytime), max(gss$Studytime)),
                     breaks = seq(min(gss$Studytime), max(gss$Studytime), 1)) +
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        # a dot plot's y axis counts stacked dots, which is not meaningful here
        panel.background = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank())
```

## Histogram

Both packages choose a number of bins for you, but they choose differently.
`hist()` uses Sturges' method, the reasonable default we discussed in class.
`geom_histogram()` just uses 30 bins and warns you to pick your own.

```{r fig.height=4, fig.width=5}
hist(gss$Studytime,
     xlab = 'Study Time (hrs)',
     ylab = 'Frequency',
     main = 'College Student Weekly Time Spent Studying')
```

```{r fig.height=4, fig.width=5}
ggplot(data = gss, aes(x = Studytime)) +
  geom_histogram(color = 'black', fill = 'lightblue') +
  theme_classic2() +
  ylab("Frequency") +
  xlab("Study Time (hrs)") +
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14))
```

To set the bins yourself, use the `bins` or `binwidth` argument — see
`?geom_histogram`. Here is the square-root rule from class:

```{r fig.height=4, fig.width=5}
n = length(gss$Studytime)
k = ceiling(sqrt(n))   # number of bins = square root of n, rounded up

ggplot(data = gss, aes(x = Studytime)) +
  geom_histogram(color = 'black', fill = 'lightblue', bins = k) +
  theme_classic2() +
  ylab("Frequency") +
  xlab("Study Time (hrs)") +
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14))
```

## Boxplot

```{r fig.height=4, fig.width=6}
boxplot(gss$Studytime,
        horizontal = T,                 # lay it on its side
        xlab = 'Study Time (Hrs)',
        main = 'College Student Weekly Study Time')
```

### Labelling the five-number summary

`text()` adds labels to any base R plot, so we can write the five-number summary
directly onto the boxplot. Compute the five numbers first:

```{r fig.height=4, fig.width=6}
minss = min(gss$Studytime)
Q1    = quantile(gss$Studytime, 0.25, type = 2)
medss = median(gss$Studytime)
Q3    = quantile(gss$Studytime, 0.75, type = 2)
maxss = max(gss$Studytime)

# the five numbers double as the x-positions of their own labels
estimates = c(minss, Q1, medss, Q3, maxss)

boxplot(gss$Studytime,
        horizontal = T,
        xlab = 'Study Time (Hrs)',
        main = 'College Student Weekly Study Time')

text(x = estimates,
     # paste() glues each name to its value: "Min=1", "Q1=5", ...
     labels = paste(c("Min", "Q1", "Median", "Q3", "Max"), estimates, sep = "="),
     # stagger the heights so neighbouring labels do not overlap
     y = c(1.3, 1.25, 1.3, 1.25, 1.3), cex = 0.6)

# the IQR goes above the box
text(x = medss, labels = paste("IQR", Q3 - Q1, sep = " = "), y = 1.5, cex = 0.7)
```

### The same thing in ggplot2

Labelling takes more work here, using `geom_label()` — one call per label, each
positioned by hand.

```{r fig.height=6, fig.width=8}
ggplot() +
  geom_boxplot(aes(x = gss$Studytime),
               fill = 'lightgrey',
               outlier.shape = 23,      # diamond
               outlier.fill = 'green',
               outlier.size = 5) +
  # mark the min and max, which the box itself does not show
  geom_point(aes(x = c(minss, maxss), y = c(0, 0)),
             shape = 21, fill = 'red', size = 4) +
  # one label per summary value; y sets how high above the axis each sits
  geom_label(aes(x = minss, y = 0.05),  label = paste('min =', minss),    label.size = 0.3) +
  geom_label(aes(x = maxss, y = 0.05),  label = paste('max =', maxss),    label.size = 0.3) +
  geom_label(aes(x = medss, y = 0.45),  label = paste('median =', medss), label.size = 0.3) +
  geom_label(aes(x = Q1,    y = 0.405), label = paste('Q1 =', Q1),        label.size = 0.3) +
  geom_label(aes(x = Q3,    y = 0.405), label = paste('Q3 =', Q3),        label.size = 0.3) +
  geom_label(aes(x = medss, y = 0.495), label = paste('IQR =', Q3 - Q1),  label.size = 0.3) +
  # leave half an hour of padding either side so the labels are not clipped
  scale_x_continuous(limits = c(minss - 0.5, maxss + 0.5), n.breaks = 10) +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.line.y = element_blank()) +
  xlab('Study Time (Hrs)') +
  ggtitle('College Student Weekly Study Time')
```
