---
title: "Homework 1 Solutions"
output:
  html_document:
    keep_md: false   ## no-op; keeps html_document a MAP so _output.yml merges in
  pdf_document: default
always_allow_html: true
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{colortbl}
  - \usepackage{wrapfig}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=8, fig.pos = "H")
#knitr::opts_chunk$set(list(echo = FALSE, eval = FALSE))
library(kableExtra)
# emit tables in the native format of whichever output is being rendered
# (hardcoding 'html' made kable dump raw HTML into the PDF, which LaTeX
#  flattened into a vertical list of cell values instead of a table)
tbl_fmt <- if (knitr::is_latex_output()) "latex" else "html"
# red-highlight answer text inside table cells, in whichever format is rendering
# (a bare <span> would be printed verbatim in the PDF)
red <- function(x) {
  if (knitr::is_latex_output()) paste0("\\textcolor{red}{", x, "}")
  else paste0('<span class="soln">', x, '</span>')
}
library(knitr)
library(latex2exp)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(plyr)
path = '../Data/'
set.seed(123)

#<span class="soln">text</span>
```

```{css, echo=FALSE}
/* Solutions are marked with <span class="soln">, not an inline style, so that
   the two rules below can reach cases an inline colour cannot. */
.soln { color: #c00000; }

/* MathJax renders $$...$$ into its own element tree, and that output does NOT
   inherit the surrounding colour -- so an answer written entirely as display
   maths inside a .soln span came out BLACK while the prose answers around it
   were red. Problem 15 (c), (d) and (g) were the visible symptom: their answers
   are nothing but display maths. Force the colour onto MathJax's own output.
   The selectors cover MathJax v2's HTML-CSS, CommonHTML and SVG outputs, since
   which one runs depends on the browser. */
.soln .MathJax, .soln .MathJax_Display, .soln .MathJax_Preview,
.soln .mjx-chtml, .soln .mjx-math, .soln mjx-container,
.soln .math, .soln .math * { color: #c00000 !important; }
.soln svg, .soln svg * { fill: #c00000 !important; }

@media print {
  /* Chrome's print-color-adjust defaults to "economy", which recolours text to
     black to save ink. Without this the whole solutions PDF prints with no red
     at all, and is indistinguishable from the blank assignment. */
  .soln, .soln * {
    color: #c00000 !important;
    -webkit-print-color-adjust: exact;
    print-color-adjust: exact;
  }
  .soln svg, .soln svg * { fill: #c00000 !important; }
}
```




${\bf (1)}$	Define the following terms: 

* **(a)** Population - the collection of all possible persons, events, or objects of interest
* **(b)** Sample - the subset of the population that we actually observe (the observed observations) 
* **(c)** Parameter - A numerical characteristic of a population (alternatively: A function of the entire population) 
* **(d)** Statistic -  A numerical characteristic of a sample (alternatively: A function of the sample) 


${\bf (2)}$ Explain the difference between a **qualitative** (categorical) and a **quantitative** variable 

<span class="soln">A qualitative variable is a non-numeric characteristic such as a name or label while a quantitative variable is numerical characteristic such as a measurement or count</span>


${\bf (3)}$ Explain the difference between a **discrete** and a **continuous** variable and give an example of each

<span class="soln">A discrete variable is a quantitative variable that takes on only distinct whole number (integer) values such as counts whereas a continuous variable is a quantitative variable that can assume any value within a certain interval such temperature or height</span>


${\bf (4)}$ Explain the difference between a **nominal** and an **ordinal** variable and give an example of each

<span class="soln">A nominal variable is a qualitative variable that has categories with no inherent ordering such as the names of a city or the brands or shoes. An ordinal variable is a qualitative variable that has an inherent ordering to its categories such as a Likert score or education level.</span>

${\bf (5)}$ **At What age did women marry?**  A historian wants to estimate the average age at marriage of women in New England in the early 19th century. Within her state archives, she finds marriage records for the years $1800 - 1820$, which she treats as a sample of all marriage records from the early 19th century. The average age of the women in the records is $24.1$ years of age. Using the appropriate statistical method, she estimates that the average age of brides in the early $19^{th}$- century New England was between $23.5$ and $24.7$ years of age. 

* **(a)** Which part of this example gives a descriptive summary of the data?
<span class="soln">When she computes a mean of $24.1$ years of age from the marriage records</span>
* **(b)** Which part of this example draws an inference about a population?
<span class="soln">The part where she estimates the average age of women to be married in the 19th century to be between $23.5$ and $24.7$ years for the population</span>
* **(c)** What population is the historian studying?
<span class="soln">All women who married in New England in the early 19th century</span>
* **(d)** The average age the historian computed from the historical records was 24.1 years of age. Is 24.1 years of age a statistic or a parameter? Why?
<span class="soln">A statistic. The quantity $24.1$ was computed from a sample of marriage records spanning the first $20$ years of the century</span>


${\bf (6)}$ Consider the following data obtained from flipping a coin 15 times. A value of $H$ denotes the coin was heads whereas a value of $T$ denotes the coin flip was tails.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
df = data.frame(`coin flip result:` = sample(c("H", "T"), 15, replace = T))

kable(t(df), format = tbl_fmt, booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))

```
Compute the frequency table for the variable "coin flip result" and answer the following questions:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
x1 = c('H', 'T')
x2 = c(7,8)
x3 = c('7/15 = 0.47','8/15 = 0.53')

kable(cbind.data.frame(x1, x2, x3), format = tbl_fmt, col.names = c('Result', 'Frequency', 'Relative Frequency'), booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))

```
* **(a)** What type of variable is "Coin Flip Result"?
<span class="soln">Qualitative nominal variable</span>
* **(b)** What proportion of the coin flips were heads?
<span class="soln">$0.47$ or $47\%$</span>
* **(c)** Why do we not compute the cumulative relative frequency for this variable?
<span class="soln">The variable is nominal and therefore has no inherent ordering for summing up the relative frequencies</span>
* **(d)** What is the best graphical display for this variable and why?
<span class="soln">Since there are only two categories (heads or tails) a pie chart or bar graph would be sufficient to show the distribution</span>



${\bf (7)}$ A survey about color preferences reported the age distribution of the people who responded. Below are the results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

ag = c('1-18','19-24','25-35','36-50','51-69','70 and over')
cts = c(10, 97, 70, 36, 14, 5)
rf = red(c('0.04', '0.42', '0.30', '0.16', '0.06', '0.02'))

age.dist = cbind.data.frame(c('Age Group (Years)', 'Counts', red('Relative Frequency')), rbind.data.frame(ag, cts, rf))

kable(age.dist, format = tbl_fmt, col.names = rep('', 7), booktabs = T, row.names = F,escape = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))

```
Use this table to answer parts a - d

* **(a)** Compute the relative frequency for each age group

* **(b)** Make a bar graph where the heights of the bars are relative frequencies
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
set.seed(123)

ag = c('1-18','19-24','25-35','36-50','51-69','70 and over')
cts = c(10, 97, 70, 36, 14, 5)

rf = cts/sum(cts)

ggplot()+geom_bar(aes(x = ag, y = rf), stat = 'identity',
                  color = 'black', fill = 'lightblue')+
  theme_bw()+
  xlab('Age Group')+
  ylab('Relative Frequency')+
  theme(axis.text = element_text(size = 12))


```
* **(c)** Describe the distribution
<span class="soln">the distribution is slightly right skew indicating that most respondents were between 19 and 35 years of age</span>
* **(d)** Explain why your bar graph is not a histogram
<span class="soln">A histogram has intervals of equal length</span>

${\bf (8)}$ Email spam is the curse of the internet. The table below gives a compilation of the most common types of spam 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

`Type of Spam` = c('Adult','Financial','Health','Leisure','Products','Scams')
Percentage = c(14.5, 16.2, 7.3, 7.8, 21.0, 14.2)

spam.dist = cbind.data.frame(`Type of Spam`, Percentage)

kable(spam.dist, format = tbl_fmt, booktabs = T, row.names = F)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))

```

Use the table to answer the questions a and b

* **(a)** Report the modal spam category
<span class="soln">The modal category is "Products" because it has the highest frequency</span>
* **(b)** Construct a **Pareto chart** using the table above
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}

ggplot(data = spam.dist, aes(x = reorder(`Type of Spam`, -Percentage), y = Percentage))+
  geom_bar(stat = 'identity', color = 'black', fill = 'lightblue2')+
  theme_hc()+
  xlab('Type of Spam')+
  ylab('Percentage')+
  ggtitle('Pareto Chart of Spam Categories')
```



${\bf (9)}$ A farmer in Idaho is interested in the number of rainy days in a given year so he records the number of rainy $R$, cloudy $C$ and sunny $S$ days over two weeks in May. His observations are $$\{R,R,C,C,C,S,S,C,R,R,C,R,S,S\}$$ Use the farmer's observations to answer the following questions.  

* **(a)** Based on the farmer's observations, what proportion of days were rainy? 
<span class="soln">$$\hat{p}_{\text{rainy}} = \frac{5}{14} = 0.36$$ </span>
* **(b)** Is the proportion in part **(a)** a statistic or a parameter? why?
<span class="soln">It is a statistic because it was computed from a sample of 14 observations of the weather</span>
* **(c)** What population is the farmer studying? is it a finite or infinite population?
<span class="soln">The population is all 365 days in a given year. It is finite</span>


${\bf (10)}$ Consider the following data from a survey conducted on college students in the state of Florida. As part of their research, surveyors recorded the high school performance (measured in grade point average - GPA) of 35 college students from across the state. For your convenience the data have been sorted from least to greatest:

$$2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,$$ 
$$3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,$$ 
$$3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,$$ 
$$3.8, 3.8, 3.8, 4.0, 4.0$$


The following frequency table gives the distribution of the variable high school GPA (with values rounded to nearest 0.5). Fill in the table and answer the following questions:

```{r, fig.cap="Frequency Table for GPA after rounding values to the nearest 0.5",  echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
#fldata = read.csv(paste0(path, 'fl_student_survey.csv'))
#fldata$gpa_rounded = round_any(fldata$high_sch_GPA, 0.5)
x_raw = c(2.0, 2.1, 2.3, 2.8, 3.0, 3.0, 3.0, 3.0, 3.0, 3.0,3.0, 3.0, 3.0, 3.1, 3.2, 3.3, 3.3, 3.4,  3.4, 3.4,3.4, 3.5, 3.5, 3.5, 3.5, 3.6, 3.6, 3.7, 3.7, 3.8,3.8, 3.8, 3.8, 4.0, 4.0)
x=round_any(x_raw, 0.5)
hsGPA_values = sort(unique(x))
hsGPA_freq = summary(as.factor(x))
hsGPA_RF = hsGPA_freq/sum(hsGPA_freq)
hsGPA_cRF = cumsum(hsGPA_RF)


FT = cbind.data.frame(GPA = hsGPA_values, 
                      Frequency = hsGPA_freq,
                      `Relative frequency` = round(hsGPA_RF,2),
                      `Cumulative RF` = round(hsGPA_cRF, 2))

kable(FT, format = tbl_fmt, digits = 2, row.names = F, booktabs = T)%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))
                      

```


* **(a)** What type of variable is GPA?
<span class="soln">Quantitative continuous</span>
* **(b)** What proportion of college students had a high school GPA $> 3.0$? (use the rounded frequency table)
<span class="soln">From the rounded frequency table, $14 + 6 = 20$ of the 35 students fall above 3.0, so the proportion is $20/35 = 0.57$. (Working from the raw, unrounded data instead gives $22/35 = 0.63$.)</span>
* **(c)** What proportion of college students had a high school GPA $< 3.0$? (use the rounded frequency table)
<span class="soln">From the rounded frequency table, $2 + 1 = 3$ of the 35 students fall below 3.0, so the proportion is $3/35 = 0.09$. (Working from the raw, unrounded data instead gives $4/35 = 0.11$.)</span>
* **(d)** What is the **mean** GPA in this sample?
<span class="soln">Using the raw data,
$$\bar{x} = \frac{2.0+2.1+...+4.0+4.0}{35} = \frac{114.5}{35} = 3.27 \approx 3.3$$
Using the rounded frequency table instead gives a very slightly different value,
$$\bar{x} = \frac{2(2.0)+1(2.5)+12(3.0)+14(3.5)+6(4.0)}{35} = \frac{115.5}{35} = 3.30$$</span>
* **(e)** What is the **median** GPA in this sample?
<span class="soln">median $= 3.4$</span>
* **(f)** What is the **mode** of GPA in this sample?

<span class="soln">Using the frequency table on the rounded values the mode is the value with the highest relative frequency $ = 3.5$</span>

<span class="soln">(Using the raw data) the mode is the most frequent value observed in the data which is $3.0$ </span>


* **(g)** Construct a dot plot for the variable GPA (hint use the frequency table)
```{r, fig.cap = "(A) Dot plot using the rounded GPA values from the frequency table (B) Dot plot using the raw GPA values", fig.height=3, fig.width=7, echo=FALSE, message = FALSE, warning=FALSE}

df = cbind.data.frame(Raw = x_raw, Rounded = x)
B = ggplot(data = df, aes(x = Rounded))+
  geom_dotplot(dotsize = 1)+
  theme_classic()+
  ylab("")+
  xlab("Rounded Student High School GPA")+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank())
A = ggplot(data = df, aes(x = Raw))+
  geom_dotplot(dotsize = 1)+
  theme_classic()+
  ylab("")+
  xlab("Sugar (g)")+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank())


ggarrange(B, A, labels = c("A", "B"), nrow = 1, ncol = 2)
  
```

* **(h)** Construct a histogram for the variable GPA (hint use the frequency table)
```{r, fig.cap = "(A) Histogram using the rounded GPA values (e.g 4 total bins) from the frequency table", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE, comment="", results="asis"}
A = ggplot(data = df, aes(x = Rounded))+
  geom_histogram(breaks = seq(2, 4, 0.5), color = 'black', fill = 'grey')+
  theme_classic()+
  ylab("Frequency")+
  xlab("Rounded Student High School GPA")+
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14))
plot(A)
```

* **(i)** Construct a stem and leaf plot for the variable GPA
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}
print("Using the Rounded GPA scores:")
print(stem(df$Rounded))
print("Using the Raw Data:")
print(stem(df$Raw, scale = 3))
```

${\bf (11)}$ Which statistic is more resistant to outliers, the mean or median? Why?

<span class="soln">The median is resistant to outliers because it relies on only the middle value (or middle two values if $n$ is even). Therefore, how far a value is from the center does not influence the median. The mean, however, is computed using \textbf{ALL} values in a sample distribution. This causes the mean to be pulled in the direction of extreme values.</span>

${\bf (12)}$ Describe the shape of the following distributions and for each distribution identify if the mean will be larger, smaller or the same as the median.

**(a)**
<span class="soln">Skew right. The mean will be greater than median because the center of the data is pulled toward the right side of the distribution</span>


```{r, fig.height=3, fig.width=4, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)
sr = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 2, 10)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sr)
```


**(b)**   
<span class="soln">Symmetric/Bell shaped. The mean and the median will be equal</span>


```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sym = ggplot()+
  geom_histogram(aes(x = rnorm(100000)), fill = 'lightblue', size = 1)+
  theme_base()+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sym)
```


**(c)**   
<span class="soln">Skew left. The mean will be less than median because the center of the data is pulled toward the left side of the distribution</span>


```{r, echo=FALSE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
set.seed(123)
sl = ggplot()+
  geom_histogram(aes(x = rbeta(100000, 10, 2)), fill = 'lightblue', size = 1)+
  theme_base()+
  # geom_vline(xintercept = mean(rbeta(100000, 10, 2)), linetype = 'dotted', color = 'red')+
  # geom_text(aes(x = mean(rbeta(100000, 10, 2)), y = 0), label = 'mean', color = 'red')+
  # geom_vline(xintercept = median(rbeta(100000, 10, 2)), linetype = 'dotted', color = 'green')+
  # geom_text(aes(x = mean(rbeta(100000, 10, 2)), y = 0), label = 'median', color = 'green')+
  theme(axis.text.x = element_text(size = 12), axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())+
  xlab("")
plot(sl)
```



\newpage
${\bf (13)}$ Consider the following four sets of observations of a quantitative variable $x$. For your convenience the observations have been sorted in increasing order. Match datasets $1-4$ with the correct histogram (labeled $A - D$)

$$\text{Dataset 1} = \{0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6\}$$
$$\text{Dataset 2}= \{0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3\}$$
$$\text{Dataset 3} = \{1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7\}$$
$$\text{Dataset 4} = \{3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7\}$$
```{r, echo=FALSE, message=FALSE, warning=F}

set.seed(123)


a = c(0.1, 1.1, 2.6, 2.7, 3.4, 3.4, 4.1, 4.4, 8.8, 9.6)
b = c(0.1, 0.3, 1.2, 2.4, 4.4, 4.5, 8.0, 8.9, 9.3, 9.3)
c = c(1.1, 3.8, 5.3, 6.0, 6.2, 6.9, 7.9, 7.9, 8.1, 8.7)
d = c(3.4, 4.5, 5.4, 5.6, 7.0, 8.5, 8.9, 9.2, 9.7, 9.7)

par(mfrow = c(2,2))
hist(a, xlab = "", breaks = seq(0, 10, 2), main = "Distribution A")
hist(c, xlab = "", breaks = seq(0, 10, 2), main = "Distribution B")
hist(d, xlab = "", breaks = seq(0, 10, 2), main = "Distribution C")
hist(b, xlab = "", breaks = seq(0, 10, 2), main = "Distribution D")
par(mfrow = c(1,1))
```


* <span class="soln">Dataset 1 goes with histogram A</span> 
* <span class="soln">Dataset 2 goes with histogram D</span>
* <span class="soln">Dataset 3 goes with histogram B</span>
* <span class="soln">Dataset 4 goes with histogram C</span>


${\bf (14)}$ Consider the following set of 10 observations of a variable $X$ sorted from least to greatest:
$$3.3, 3.8, 4.0, 4.8, 4.8, 5.1, 5.2, 5.6, 5.7, 6.9$$
Use the data to answer parts a-b

* **(a)** Given the data, compute the $40^{th}$ percentile of $X$
<span class="soln">The $p^{th}$ percentile is the value with $p\%$ of the observations **at or below** it, so build up the cumulative relative frequency and find where it passes $0.40$: $CRF(3.3) = 0.10$, $CRF(3.8) = 0.20$, $CRF(4.0) = 0.30$, then $CRF(4.8) = 0.50$ because $4.8$ occurs twice.
No value has a cumulative relative frequency of exactly $0.40$ - it jumps straight from $0.30$ at $4.0$ to $0.50$ at $4.8$ - so the $40^{th}$ percentile is the value at that jump:
$$P_{40} = 4.8$$
Checking against the definition: only $3$ of the $10$ observations ($30\%$) are at or below $4.0$, while $5$ of them ($50\%$) are at or below $4.8$, so $4.8$ is the smallest value with at least $40\%$ of the data at or below it.</span>
* **(b)** Compute Interquartile Range (IQR) range of $X$
<span class="soln">Use the **method of halves**. Here $n = 10$ is even, so there is no middle value to leave out - the data splits straight down the middle into two halves of five:
$$\underbrace{3.3, \ 3.8, \ 4.0, \ 4.8, \ 4.8}_{\text{lower half}} \qquad \underbrace{5.1, \ 5.2, \ 5.6, \ 5.7, \ 6.9}_{\text{upper half}}$$
$Q1$ is the median of the lower half and $Q3$ is the median of the upper half. Each half has five values, so each median is just its middle (third) value:
$$Q1 = 4.0 \qquad Q3 = 5.6$$
$$IQR = Q3 - Q1 = 5.6 - 4.0 = 1.6$$</span>


${\bf (15)}$ Consider the following $n = 20$ observations of several popular cereal brands and answer questions a - g about the variable "Sugar (g)":

```{r, echo = FALSE, message=FALSE, warning=FALSE}
cereal = read.csv(paste0(path, 'cereal.csv'))
kable(cereal, format = tbl_fmt, col.names = c("Brand", "Sodium (mg)", "Sugar (g)", 'Type'))%>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, width = "18em")%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))

```

* **(a)** Construct a histogram of the data using the following bins: $\leq 2 (g)$, $2 - 4 (g)$, 
$4 - 6 (g)$, $6 - 8 (g)$,$8 - 10 (g)$, $10 - 12 (g)$, $12 - 14 (g)$,$14 - 16 (g)$,$> 16 (g)$ (hint start by creating a frequency table) 

```{r, fig.cap = "(A) Histogram of the variable sugar (g) using the bins in the frequency table (above)", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE}
A = ggplot(data = cereal, aes(x = Sugar))+
  geom_histogram(breaks = seq(0, 20, 2), color = 'black', fill = 'grey')+
  theme_classic()+
  ylab("Frequency")+
  xlab("Sugar (g)")+
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14))
plot(A)

x_rounded = hist(cereal$Sugar, breaks = seq(0, 20, 2), plot = F)
x_rf = x_rounded$counts/sum(x_rounded$counts)
x_crf = cumsum(x_rf)

Tabular = cbind.data.frame(`Sugar (g)` = c("< = 2 (g)", 
                                           "2 - 4 (g)",
                                           "4 - 6 (g)",
                                           "6 - 8 (g)",
                                           "8 - 10 (g)",
                                           "10 - 12 (g)",
                                           "12 - 14 (g)",
                                           "14 - 16 (g)",
                                           "> 16 (g)"),
                           Frequency = x_rounded$counts[1:9],
                           `RF(x)` = x_rf[1:9],
                           `CRF(x)` = x_crf[1:9])

kable(Tabular, format = tbl_fmt, digits = 2, booktabs = T,
      caption = "Frequency table used to construct histogram for the variable \"Sugar (g)\"")%>%
  kable_classic(full_width = F, html_font = "Cambria")%>%kable_styling(bootstrap_options = 'striped', latex_options = c('striped', 'HOLD_position'))
  
```


<br>

* **(b)** Compute the mean, median, and mode of the variable "Sugar (g)"
<span class="soln">Using the raw data there are 5 modes each with a frequency of 2
mode(s) $= 3,4,11,12,14$</span>
<span class="soln">median $= \frac{9+10}{2} = 9.5$</span>
<span class="soln">$$\bar{x} = \frac{1}{20}\sum_{i=1}^{20} x_i = \frac{(11+18+5+14+...+3+11)}{20} = 8.75$$ (note that the mean may be slightly different if you used the frequency table above</span>

* **(c)** Compute the variance and standard deviation of the variable "Sugar (g)"
<span class="soln">$$s^2 = \frac{1}{19}\sum_{i=1}^{20} (x_i-\bar{x})^2 = \frac{(11 - 8.75)^2 + (18-8.75)^2 + ...+(3-8.75)^2 + (11 - 8.75)^2}{19}$$  $$ = \frac{5.06 + 85.56 + ...+33.06+5.06}{19} = 28.3$$ $$s = \sqrt{s^2} = \sqrt{28.3} = 5.32$$</span>


* **(d)** Compute the quartiles (Q1, Q2, Q3) and interquartile range of the variable "Sugar (g)"
<span class="soln">$$Q1 = 4, \ Q2 = \text{median} = 9.5, \ Q3 = 13, \ IQR = 13 - 4 = 9$$</span>

* **(e)** Create a box and whisker plot for the variable "Sugar (g)" and mark the Q1, Q2, and Q3 quartiles, the median and any potential outliers

```{r, fig.cap = "Boxplot of Sugar content in 20 cereal brands", fig.height=5, fig.width=6, echo=FALSE, message = FALSE, warning=FALSE}
X = cereal$Sugar

mi = min(X)
mx = max(X)
Q1 = (4+4)/2
med = 9.5
Q3 = (12+14)/2
estims = c(mi, Q1, med, Q3, mx)
boxplot(X, horizontal = TRUE, axes = TRUE, staplewex = 1)
text(x=estims-0.1, labels =paste(c("Min", "Q1","Median", "Q3", "Max"), 
                                   estims, sep = "="), y=1.25, cex = 0.68)
text(x=Q3-Q1, labels =paste("IQR", Q3 - Q1, sep = " = "), y=1.5, cex = 0.68)
```

* **(f)** Plot the Cumulative Distribution of the variable "Sugar (g)"

```{r, fig.cap = "Cumulative Distribution of Sugar (g)", fig.height=3, fig.width=4, echo=FALSE, message = FALSE, warning=FALSE}

A=ggplot()+geom_step(aes(x = x_rounded$breaks[2:10], y = x_crf[1:9]), size = 0.8)+
  geom_point(aes(x = x_rounded$breaks[2:10], y = x_crf[1:9]), size = 3)+
  theme_hc()+
  xlab('Sugar (g)')+
  ylab('Cumulative Relative Frequency')+
  scale_x_continuous(breaks = seq(0, 20, 2))+
  scale_y_continuous(breaks = seq(0, 1, 0.1))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))
  
  plot(A)
```

* **(g)** What proportion of cereals have a sugar content less than or equal to 10 grams?
<span class="soln">$$55\%$$</span>



<!-- **(16)** True or False: The range is the best descriptive statistic for the spread of a distribution (Explain your answer) -->

<!-- **(17)** True or False: The standard deviation $s$ is the square of the variance -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->

<!-- **(18)** True or False: A **deviation} is defined as the most frequent value in a distribution -->
<!-- * **(a)** True -->
<!-- * **(b)** False -->












